---
title: "BLUP Accuracy in Multi-Environment Trials"
subtitle: "`accuracy()` and `plot_accuracy()`"
author: "Julian Taylor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
asreml_ok <- requireNamespace("asreml", quietly = TRUE)
library(biomAid)
library(ggplot2)
if (asreml_ok) {
  library(asreml)
  asreml.options(trace = FALSE)
}
```

---

## Overview

`accuracy()` computes model-based BLUP accuracy for each environment (or group)
in a multi-environment trial (MET) fitted with ASReml-R V4. Two complementary
metrics are available:

- **Mrode accuracy** — per-variety accuracy derived from the prediction error
  variance (PEV) and the genetic variance.
- **Cullis H²** — a reliability analogue based on the mean squared standard
  error of difference (SED) between BLUPs.

Both metrics are bounded between 0 and 1, with values closer to 1 indicating
better prediction. The function handles all major ASReml-R V4 random
structures for the variety term:
`fa()`, `diag()`, `corgh()`, `corh()`, `us()`, and single-environment `id()`.

`plot_accuracy()` visualises these results with six complementary plot types
covering single-model summaries and head-to-head model comparisons.

---

## Mathematical Framework

### Genetic variance and BLUP prediction error

Let $\hat{u}_{ij}$ be the BLUP for variety $i$ in environment $j$, with
prediction error variance

$$
\operatorname{PEV}_{ij}
  = \operatorname{Var}(\hat{u}_{ij} - u_{ij}).
$$

The genetic (co)variance for environment $j$ is denoted $G_{jj}$ — the
$j$-th diagonal of the **G-matrix** (or the corresponding variance component
for simpler structures such as `diag()`).

---

### Mrode accuracy

Following Mrode (2014), the accuracy for variety $i$ in environment $j$ is

$$
r_{ij} = \sqrt{1 - \frac{\operatorname{PEV}_{ij}}{G_{jj}}},
$$

which is the square root of the **reliability** — the proportion of true
genetic variance that is recovered by the BLUP. The **mean accuracy** for
environment $j$ is

$$
\bar{r}_j = \frac{1}{n_j}\sum_{i=1}^{n_j} r_{ij},
$$

where $n_j$ is the number of varieties with observed data in environment $j$.

`accuracy()` invokes the argument `only =` in `predict.asreml()` to ensure the correct
$\operatorname{PEV}_{ij}$ are returned for the target random term supplied.

---

### Generalised H²

Cullis et al. (2006) proposed a heritability analogue based on the mean
pairwise squared standard error of difference (SED):

$$
H^2_j = 1 - \frac{\overline{\text{SED}}^2_j}{2\,G_{jj}},
$$

where $\overline{\text{SED}}^2_j$ is the mean of the upper-triangular
elements of the SED matrix for environment $j$. Unlike Mrode accuracy,
$H^2_j$ captures the precision of **variety comparisons** rather than
individual predictions, and is equivalent to heritability on a variety-mean
basis.

---

### Extracting $G_{jj}$ from ASReml-R V4

The extraction strategy depends on the random structure fitted for the
variety term:

| Structure | $G_{jj}$ | `only=` string |
|-----------|-----------|----------------|
| `fa(Env, k):id(Var)` | $\sum_f \lambda_{jf}^2 + \psi_j$ | `"fa(Env, k):Var"` |
| `diag(Env):id(Var)` | $\sigma^2_j$ | `"Env:Var"` |
| `corgh(Env):id(Var)` | $\sigma^2_j$ | `"Env:Var"` |
| `corh(Env):id(Var)` | $\sigma^2_j$ | `"Env:Var"` |
| `us(Env):id(Var)` | $\sigma^2_{jj}$ (diagonal of **U**) | `"Env:Var"` |
| `id(Var)` | $\sigma^2$ | `"Var"` |

For factor analytic models, $G_{jj} = \boldsymbol{\lambda}_j^\top
\boldsymbol{\lambda}_j + \psi_j$ where $\boldsymbol{\lambda}_j$ is the
$k$-vector of factor loadings for environment $j$ and $\psi_j$ is the
specific variance.

---

## Setup

### Simulated data

> **Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
> needed to read it; run the code chunks only if you have the package
> installed.

We use `simTrialData()` to generate a MET with 50 varieties across 8
environments, 2 replicates, and an `"unbalanced"` incidence structure. The
data-generating model uses `n_fa = 2` latent factors so environments share
genuine genetic covariance, and the incidence ensures that varieties appear
in different subsets of environments — producing meaningful variety-to-variety
variation in prediction error variance that makes the per-variety accuracy
plots informative.

```{r simulate}
#| cache: true
#| eval: !expr asreml_ok
out <- simTrialData(
  nvar      = 50L,
  nsite     = 8L,
  n_fa      = 2L,
  nrep      = 2L,
  incidence = "unbalanced",
  seed      = 11L,
  verbose   = FALSE
)
dat         <- out$data
dat$Row     <- factor(dat$Row)
dat$Column  <- factor(dat$Column)
dat$Site    <- factor(dat$Site)
dat$Variety <- factor(dat$Variety)

# Connectivity summary
cat("Varieties per site: mean =", round(mean(colSums(out$params$incidence)), 1),
    " range [", min(colSums(out$params$incidence)), ",",
    max(colSums(out$params$incidence)), "]\n")
cat("Sites per variety:  mean =", round(mean(rowSums(out$params$incidence)), 1),
    " range [", min(rowSums(out$params$incidence)), ",",
    max(rowSums(out$params$incidence)), "]\n")
```

A quick look at the data structure:

```{r show_data}
#| eval: !expr asreml_ok
head(dat[, c("Site", "Variety", "Rep", "Row", "Column", "yield")])
```

---

### Fitting three models

We fit three random structures for the Site × Variety genetic term, in
increasing order of complexity:

| Model | Variety random structure | Genetic parameters |
|-------|-------------------------|--------------------|
| **Diag** | `diag(Site):id(Variety)` | 6 (one $\sigma^2_j$ per site) |
| **FA(1)** | `fa(Site, 1):id(Variety)` | 12 (6 loadings + 6 specific variances) |
| **FA(2)** | `fa(Site, 2):id(Variety)` | 18 (12 loadings + 6 specific variances) |

All three models include `at(Site):Rep` for replicate effects and a
common homogeneous residual. The `diag` model assumes no genetic correlation
between sites — varieties are re-ranked completely independently across
environments. The FA models borrow information across environments: FA(1)
captures one dominant axis of GEI; FA(2) adds a second, allowing richer
rank-change patterns.

```{r fit_diag}
#| cache: true
#| dependson: simulate
#| eval: !expr asreml_ok
#| results: hide
m_diag <- asreml(
  yield    ~ Site,
  random   = ~ diag(Site):id(Variety) + at(Site):Rep,
  data     = dat,
  maxit    = 30L
)
m_diag <- update(m_diag)
```

```{r fit_fa1}
#| cache: true
#| dependson: fit_diag
#| eval: !expr asreml_ok
#| results: hide
m_fa1 <- asreml(
  yield    ~ Site,
  random   = ~ fa(Site, 1):id(Variety) + at(Site):Rep,
  data     = dat,
  maxit    = 30L
)
m_fa1 <- update(m_fa1)
```

```{r fit_fa2}
#| cache: true
#| dependson: fit_fa1
#| eval: !expr asreml_ok
#| results: hide
m_fa2 <- asreml(
  yield    ~ Site,
  random   = ~ fa(Site, 2):id(Variety) + at(Site):Rep,
  data     = dat,
  maxit    = 30L
)
m_fa2 <- update(m_fa2)
```

A quick comparison of log-likelihoods confirms the FA models provide a
better fit to the data. AIC is computed as $-2 \ell + 2p$ where $p$ is
the number of variance parameters:

```{r aic_compare}
#| eval: !expr asreml_ok
.aic <- function(m) -2 * m$loglik + 2 * nrow(summary(m)$varcomp)
data.frame(
  Model  = c("Diag", "FA(1)", "FA(2)"),
  LogLik = round(c(m_diag$loglik, m_fa1$loglik, m_fa2$loglik), 2),
  AIC    = round(c(.aic(m_diag),  .aic(m_fa1),  .aic(m_fa2)),  2)
)
```

---

## Computing accuracy with `accuracy()`

### Group-level summaries

The default call returns both Mrode accuracy and Cullis H² for every site.
Results are at the group level — one row per environment.

```{r acc_diag}
#| eval: !expr asreml_ok
acc_diag <- accuracy(m_diag, metric = c("accuracy", "gen.H2"))
acc_diag
```

```{r acc_fa1}
#| eval: !expr asreml_ok
acc_fa1 <- accuracy(m_fa1, metric = c("accuracy", "gen.H2"))
acc_fa1
```

```{r acc_fa2}
#| eval: !expr asreml_ok
acc_fa2 <- accuracy(m_fa2, metric = c("accuracy", "gen.H2"))
acc_fa2
```

The `G_jj` column reports the environment-specific genetic variance used as
the denominator. `mean_acc` is $\bar{r}_j$ and `sd_acc` captures the spread
of per-variety accuracies within the environment. `gen.H2` is $H^2_j$.

### Requesting a single metric

Pass a single string to `metric` when you only need one measure:

```{r acc_single_metric}
#| eval: !expr asreml_ok
# Mrode accuracy only
accuracy(m_fa2, metric = "accuracy")

# Cullis H² (gen.H2) only
accuracy(m_fa2, metric = "gen.H2")
```

### Per-variety accuracies

Set `by_variety = TRUE` to return one row per variety × environment. The
`present` argument controls which combinations are included:

- **`present = TRUE`** (default) — restrict to variety × environment
  combinations *observed* in the data. Produces holes in heatmaps for
  unbalanced trials.
- **`present = FALSE`** — include *all* combinations returned by
  `predict.asreml()`. For FA models the G-matrix supports prediction of
  BLUPs for unobserved combinations; their PEV is higher (lower accuracy)
  than for observed ones, reflecting reduced information.

```{r acc_bv}
#| eval: !expr asreml_ok
# present = FALSE fills the full variety x site grid for FA models
acc_fa2_bv    <- accuracy(m_fa2,  metric = c("accuracy", "gen.H2"),
                           by_variety = TRUE, present = FALSE)
acc_diag_bv   <- accuracy(m_diag, metric = c("accuracy", "gen.H2"),
                           by_variety = TRUE, present = FALSE)

head(acc_fa2_bv)
```

---

## Visualising with `plot_accuracy()`

### Lollipop — per-site accuracy summary

The lollipop plot is the go-to summary for group-level results. Each site
appears as a row, with the lollipop head at the mean accuracy. Horizontal
error bars span ±1 standard deviation of the per-variety accuracies (Mrode
metric only). When both metrics are requested the plot facets into two panels.

```{r lollipop_fa2}
#| eval: !expr asreml_ok
#| fig-height: 4.5
plot_accuracy(acc_fa2, type = "lollipop",
              metric = c("accuracy", "gen.H2"))
```

A single-metric call gives a cleaner, unfaceted plot:

```{r lollipop_fa2_acc}
#| eval: !expr asreml_ok
#| fig-height: 4
plot_accuracy(acc_fa2, type = "lollipop", metric = "accuracy")
```

Note that Cullis H² is systematically lower than Mrode accuracy —
$H^2$ measures comparison precision while $r$ measures individual prediction
reliability, and the two capture subtly different aspects of model quality.

---

### Violin — per-variety accuracy distribution

The violin plot requires `by_variety = TRUE` data and displays the Mrode
Accuracy metric only. Cullis H² is a group-level constant and is suppressed
automatically — use `type = "lollipop"` to visualise it. The width of each
violin reflects the spread of per-variety accuracies within each environment.

```{r violin_fa2}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_accuracy(acc_fa2_bv, type = "violin", metric = "accuracy")
```

Varieties that appear in fewer sites share less information with the rest of
the trial — their BLUPs have higher PEV and lower accuracy. "Core" varieties
(present in most environments) sit at the high-accuracy end; rarely-observed
varieties sit at the low end. With `present = FALSE`, the full variety × site
grid is included so unobserved combinations also contribute, showing an
additional tail of lower accuracies.

---

### Heatmap — variety × site accuracy grid

The heatmap visualises per-variety accuracy across all environments
simultaneously using `present = FALSE` data, so the full variety × site grid
is displayed without holes. Varieties are sorted from highest mean accuracy
(top) to lowest (bottom); environments are ordered by increasing mean accuracy.
The viridis colour scale maps accuracy onto a perceptually uniform gradient.

```{r heatmap_fa2}
#| eval: !expr asreml_ok
#| fig-height: 6
#| fig-width: 7
plot_accuracy(acc_fa2_bv, type = "heatmap", metric = "accuracy")
```

A column that is uniformly darker than its neighbours indicates an environment
where all varieties are predicted less reliably. Within each column, darker
tiles correspond to varieties with fewer site appearances — their BLUPs rely
more on the FA G-matrix and less on direct observations, so their PEV is
higher.

---

## Comparing models with `res2 =`

The three comparison plot types take a second accuracy object via `res2` and
directly show the change in accuracy between models. This is where the story
of moving from `diag` to FA models becomes clearly visible.

### Dumbbell — paired group-level comparison

The dumbbell plot places two dots per site connected by a horizontal segment.
The segment colour encodes the direction of change: **green** when the second
model improves accuracy, **red** when it declines.

#### Diag → FA(1)

```{r dumbbell_diag_fa1}
#| eval: !expr asreml_ok
#| fig-height: 4.5
plot_accuracy(acc_diag, acc_fa1,
              type   = "dumbbell",
              metric = "accuracy",
              label1 = "Diag",
              label2 = "FA(1)")
```

#### FA(1) → FA(2)

```{r dumbbell_fa1_fa2}
#| eval: !expr asreml_ok
#| fig-height: 4.5
plot_accuracy(acc_fa1, acc_fa2,
              type   = "dumbbell",
              metric = "accuracy",
              label1 = "FA(1)",
              label2 = "FA(2)")
```

With both metrics faceted:

```{r dumbbell_both_metrics}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_accuracy(acc_diag, acc_fa2,
              type   = "dumbbell",
              metric = c("accuracy", "gen.H2"),
              label1 = "Diag",
              label2 = "FA(2)")
```

The improvement in FA(2) over `diag` tends to be larger for environments
with stronger GEI — those are the environments where borrowing information
via the factor structure is most beneficial.

---

### Scatter — per-variety comparison

The scatter plot places each variety × environment combination as a point,
with model1 accuracy on the x-axis and model2 accuracy on the y-axis.
Points above the dashed diagonal (x = y) represent variety × environment
combinations where model2 is more accurate. Points are coloured by
environment.

```{r scatter_diag_fa2}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_accuracy(acc_diag_bv, acc_fa2_bv,
              type   = "scatter",
              metric = "accuracy",
              label1 = "Diag",
              label2 = "FA(2)")
```

The scatter plot reveals whether the accuracy gain is uniform across the
genetic material or concentrated in particular varieties — for instance, rare
or poorly-replicated varieties often benefit most from the additional
shrinkage imposed by the FA structure.

---

### Diff — signed accuracy gain

The diff plot shows (model2 − model1) for each environment as a lollipop
centred on zero. It distils the dumbbell into a single signed quantity,
making it easy to rank environments by how much they benefit from the more
complex model. The x-axis is always symmetric around zero.

#### Diag → FA(2)

```{r diff_diag_fa2}
#| eval: !expr asreml_ok
#| fig-height: 4.5
plot_accuracy(acc_diag, acc_fa2,
              type   = "diff",
              metric = "accuracy",
              label1 = "Diag",
              label2 = "FA(2)")
```

#### FA(1) → FA(2), both metrics

```{r diff_fa1_fa2_both}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_accuracy(acc_fa1, acc_fa2,
              type   = "diff",
              metric = c("accuracy", "gen.H2"),
              label1 = "FA(1)",
              label2 = "FA(2)")
```

---

## Extracting plot data — `return_data = TRUE`

All six plot types support `return_data = TRUE`, returning a named list with
`$plot` and `$data`. This is useful for further customisation or downstream
analysis.

```{r return_data}
#| eval: !expr asreml_ok
rd <- plot_accuracy(acc_diag, acc_fa2,
                    type        = "diff",
                    metric      = c("accuracy", "gen.H2"),
                    label1      = "Diag",
                    label2      = "FA(2)",
                    return_data = TRUE)

head(rd$data)
```

The ggplot object is fully extensible with standard `ggplot2` additions:

```{r bespoke}
#| eval: !expr asreml_ok
#| fig-height: 4.5
rd$plot +
  ggplot2::ggtitle(
    "Accuracy gain: FA(2) over Diag",
    subtitle = "Positive values indicate FA(2) is more accurate"
  )
```

---

## Summary

### `accuracy()` arguments

| Argument | Default | Description |
|----------|---------|-------------|
| `model` | — | Fitted ASReml-R V4 model object |
| `classify` | Auto-detected | Override the `predict()` classify string |
| `metric` | `c("accuracy","gen.H2")` | Metrics to compute; one or both |
| `pworkspace` | `"2gb"` | Passed to `predict.asreml()` |
| `by_variety` | `FALSE` | Return one row per variety × group |
| `present` | `TRUE` | `TRUE` restricts to observed variety × group combinations; `FALSE` includes all `predict.asreml()` rows (FA models support unobserved combos via the G-matrix) |

### `plot_accuracy()` types

| Type | Input | Best for |
|------|-------|---------|
| `"lollipop"` | Group-level | Compact accuracy summary per site |
| `"violin"` | `by_variety` | Distribution of accuracy across varieties |
| `"heatmap"` | `by_variety` | Variety × site accuracy at a glance |
| `"dumbbell"` | Group-level + `res2` | Paired model comparison per site |
| `"scatter"` | `by_variety` + `res2` | Per-variety model comparison |
| `"diff"` | Group-level + `res2` | Signed accuracy gain per site |

Faceting into two panels is automatic when `metric = c("accuracy","gen.H2")`.

---

## References

Cullis, B.R., Smith, A.B. & Coombes, N.E. (2006). On the design of early
generation variety trials with correlated data. *Journal of Agricultural,
Biological and Environmental Statistics*, **11**(4), 381–393.
<https://doi.org/10.1198/108571106X154443>

Mrode, R.A. (2014). *Linear Models for the Prediction of Animal Breeding
Values*, 3rd edition. CAB International, Wallingford.

Smith, A., Cullis, B. & Thompson, R. (2001). Analysing variety by environment
data using multiplicative mixed models and adjustments for spatial field
trend. *Biometrics*, **57**(4), 1138–1147.
<https://doi.org/10.1111/j.0006-341X.2001.01138.x>

Gilmour, A.R., Gogel, B.J., Cullis, B.R. & Thompson, R. (2009).
*ASReml User Guide Release 3.0*. VSN International, Hemel Hempstead.
