---
title: "Multiple Comparison Criteria for Predicted Values"
subtitle: "`compare()` and `plot_compare()`"
author: "biomAid package"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
library(asreml)
```

---

## Overview

`compare()` computes a pairwise comparison criterion — **HSD**, **LSD**, or
**Bonferroni-corrected LSD** — for predicted values obtained from an ASReml-R
V4 model.  The criterion is a single number: two predicted values in the same
group are declared significantly different if and only if their absolute
difference exceeds it.

Three comparison types are available:

| Type | Formula | Multiple-testing control | Best used when |
|------|---------|--------------------------|----------------|
| `"HSD"` (Tukey) | $\frac{\overline{\text{SED}}}{\sqrt{2}} \times q_{1-\alpha}(n_g, \nu)$ | Family-wise (exact for balanced designs) | Routine variety/treatment comparisons |
| `"LSD"` | $\overline{\text{SED}} \times t_{\alpha/2,\,\nu}$ | None | Exploratory or preplanned single comparisons |
| `"Bonferroni"` | $\overline{\text{SED}} \times t_{\alpha/(2m),\,\nu}$ | Family-wise (conservative) | Small number of specific comparisons |

The companion function `plot_compare()` turns a `compare()` result directly
into publication-ready visualisations: a sorted **dot plot** with a
criterion-width band, a **compact letter display** (CLD), or a
**pairwise significance heatmap**.

---

## Mathematical Framework

### Prediction error variance and the average SED

Let $\hat{\boldsymbol{\tau}}_g = (\hat{\tau}_{1g}, \ldots, \hat{\tau}_{n_g g})^\top$
be the vector of $n_g$ predicted values in group $g$, with prediction error
variance–covariance matrix

$$
\mathbf{V}_g = \operatorname{Var}(\hat{\boldsymbol{\tau}}_g - \boldsymbol{\tau}_g),
$$

returned by `predict(model, classify = term, vcov = TRUE)$vcov` in ASReml-R V4.

The standard error of a difference (SED) for the pair $(i, j)$ is

$$
\text{SED}_{ij} = \sqrt{V_{ii} + V_{jj} - 2V_{ij}}.
$$

`compare()` uses the **arithmetic mean SED** across all $m = \binom{n_g}{2}$
pairs as the representative spread of prediction error:

$$
\overline{\text{SED}}_g
  = \frac{1}{m}\sum_{i < j} \text{SED}_{ij}.
$$

Using the average SED (rather than a single representative pair) is robust when
PEVs vary across units — as they do in unbalanced designs and spatial models.

---

### Comparison criteria

Let $\nu = \texttt{model\$nedf}$ be the error degrees of freedom.

**Tukey HSD** — controls the family-wise error rate (FWER) exactly for
balanced designs by referring to the Studentised range distribution:

$$
\text{HSD}_g
  = \frac{\overline{\text{SED}}_g}{\sqrt{2}} \times
    q_{1-\alpha}(n_g, \nu),
$$

where $q_{1-\alpha}(n_g, \nu)$ is the upper $\alpha$ critical value of the
Studentised range with $n_g$ means and $\nu$ degrees of freedom.

**LSD** — no multiplicity correction; equivalent to a two-sided $t$-test at
level $\alpha$ for each pair:

$$
\text{LSD}_g = \overline{\text{SED}}_g \times t_{\alpha/2,\,\nu}.
$$

**Bonferroni** — applies the Bonferroni correction across all $m$ pairs,
controlling the FWER conservatively:

$$
\text{Bonferroni}_g
  = \overline{\text{SED}}_g \times t_{\alpha / (2m),\,\nu}.
$$

For all three types: $|\hat{\tau}_i - \hat{\tau}_j| > \text{criterion}$
declares pair $(i, j)$ significantly different.

---

### Posterior variance (`pev = FALSE`)

When `term` is in the **random** part of the model, the PEV
$\mathbf{V}_{\hat{\boldsymbol{\tau}}}$ underestimates the uncertainty of the
difference between BLUPs.  Setting `pev = FALSE` substitutes the **posterior
variance**

$$
\operatorname{Var}(\hat{\boldsymbol{\tau}} \mid \mathbf{y})
  = \mathbf{G} - \mathbf{V}_{\hat{\boldsymbol{\tau}}}
$$

which accounts for the shrinkage in BLUP estimation and gives a more
appropriate measure of uncertainty for genotypic BLUPs.

---

## Setup

### Simulated data and model

> **Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
> needed to read it; run the code chunks only if you have ASReml-R installed.

We use `simTrialData()` to generate a balanced multi-environment trial with
20 varieties, 4 sites, and 3 nitrogen treatments (T0, T1, T2).

```{r simulate}
#| cache: true
out <- simTrialData(
  nvar          = 20L,
  nsite         = 4L,
  treatments    = c("T0", "T1", "T2"),
  treat_effects = c(0, 80, 200),
  nrep          = 3L,
  seed          = 42L,
  verbose       = FALSE
)
dat        <- out$data
dat$Row    <- factor(dat$Row)
dat$Column <- factor(dat$Column)
```

We fit a **fixed `Treatment × Site × Variety` model** with
site-within-replicate random effects and a spatial residual structure.
Variety must be in the fixed effects so that `predict()` returns BLUEs for
every Variety × Site combination — the basis for the comparison criteria.

```{r fit_model}
#| cache: true
#| results: hide
asreml.options(trace = FALSE)
model <- asreml(
  yield    ~ Treatment * Site * Variety,
  random   = ~ at(Site):Rep,
  residual = ~ dsum(~ ar1(Row):ar1(Column) | Site),
  data     = dat,
  maxit    = 30L
)
model <- update(model)
```

---

## Example 1: HSD — single global group

The simplest use: compare all variety BLUEs across the trial in a single
global group (no `by` argument).  `compare()` calls `predict()` internally
using `term` as the `classify`.

```{r hsd_single}
res1 <- compare(
  model,
  term = "Variety",
  type = "HSD"
)

head(res1[order(res1$predicted.value, decreasing = TRUE), ])
```

The `HSD` column is **constant within the group** — all predicted values share
the same criterion.  Two varieties are significantly different if
`|pred_i - pred_j| > HSD`.  The `avsed` column records the average SED on
which the criterion is based.

Applying the criterion manually:

```{r hsd_single_apply}
# Are Var01 and Var20 significantly different?
preds <- setNames(res1$predicted.value, as.character(res1$Variety))
diff  <- abs(preds["Var01"] - preds["Var20"])
cat("|Var01 - Var20| =", round(diff, 1L),
    "  HSD =",           round(res1$HSD[1L], 1L),
    "  Significant:",    diff > res1$HSD[1L], "\n")
```

---

## Example 2: HSD by site

Adding `by = "Site"` performs independent comparisons within each site, using
each site's own prediction error variance matrix and thus its own criterion.
This is the most common usage: **20 varieties compared within each of 4 sites**.

```{r hsd_by_site}
#| cache: true
res_hsd <- compare(
  model,
  term = "Variety:Site",
  by   = "Site",
  type = "HSD"
)

# HSD varies by site (different PEV per site from the spatial model)
unique(res_hsd[, c("Site", "HSD", "avsed")])
```

The HSD differs between sites because the spatial AR1×AR1 residual structure
gives each site its own effective error variance.  Sites with tighter spatial
control (lower `avsed`) have a smaller, more powerful criterion.

---

## Example 3: Multi-factor `by`

When predictions span multiple factors, `by` accepts a character vector.
Here varieties are compared **within every Site × Treatment combination** —
asking "which varieties perform best under each treatment at each site?".
The colon-string form `"Site:Treatment"` is equivalent to the vector form.

```{r hsd_multi_by}
#| cache: true
# Compare varieties within each Site × Treatment combination
res3a <- compare(model, term = "Variety:Site:Treatment",
                 by = c("Site", "Treatment"), type = "HSD")
res3b <- compare(model, term = "Variety:Site:Treatment",
                 by = "Site:Treatment", type = "HSD")

# HSD per Site × Treatment group (first 6 of 12 groups)
unique(res3a[, c("Site", "Treatment", "HSD", "avsed")])[1:6, ]
```

The criterion varies across Site × Treatment combinations because each
sub-environment has its own spatial error structure.

---

## Example 4: LSD and Bonferroni

The three criteria differ only in the multiplier applied to `avsed`.
LSD is the most liberal (no correction); Bonferroni is the most conservative;
HSD sits between them for most group sizes.

```{r three_criteria}
#| cache: true
res_lsd <- compare(model, term = "Variety:Site", by = "Site", type = "LSD")
res_bon <- compare(model, term = "Variety:Site", by = "Site",
                   type = "Bonferroni")

# Criteria for Env1 (res_hsd already computed in Example 2)
env1 <- function(r, col) round(unique(r[[col]][r$Site == "Env1"]), 1L)
cat("Env1 criteria (20 varieties):\n")
cat("  LSD        =", env1(res_lsd, "LSD"),        "\n")
cat("  HSD        =", env1(res_hsd, "HSD"),        "\n")
cat("  Bonferroni =", env1(res_bon, "Bonferroni"), "\n")
```

For 20 varieties, $m = \binom{20}{2} = 190$ pairs.  The Bonferroni correction
is very conservative at this scale — HSD is generally preferred for large
variety trials.

---

## Example 5: Variety BLUPs with `pev = FALSE`

When variety is in the **random** effects the PEV understates uncertainty
because it ignores the shrinkage towards zero inherent in BLUP estimation.
Setting `pev = FALSE` substitutes the posterior variance
$\mathbf{G} - \mathbf{PEV}$, giving a more appropriate — and larger — criterion.

We refit the model with **Variety in the random effects** and
`Treatment × Site` fixed, so `predict()` returns BLUPs rather than BLUEs.

```{r pev_false}
#| cache: true
#| results: hide
model_rnd <- asreml(
  yield    ~ Treatment * Site,
  random   = ~ Variety + at(Site):Rep,
  residual = ~ dsum(~ ar1(Row):ar1(Column) | Site),
  data     = dat,
  maxit    = 30L
)
model_rnd <- update(model_rnd)
```

```{r pev_false_compare}
res_pev  <- compare(model_rnd, term = "Variety", type = "HSD", pev = TRUE)
res_post <- compare(model_rnd, term = "Variety", type = "HSD", pev = FALSE)

cat("HSD (PEV):                    ", round(res_pev$HSD[1L],  1L), "\n")
cat("HSD (posterior variance):     ", round(res_post$HSD[1L], 1L), "\n")
cat("Ratio (posterior / PEV HSD):  ",
    round(res_post$HSD[1L] / res_pev$HSD[1L], 2L), "\n")
```

The posterior-variance HSD is larger than the PEV-based HSD, reflecting the
additional uncertainty that BLUP shrinkage introduces.  Using `pev = TRUE`
on a random-effects term underestimates the true uncertainty of a difference
between BLUPs and will declare too many pairs significant.

> **Note:** `pev = FALSE` is only meaningful when `term` is in the **random**
> part of the model.  When the term is fixed (as in Examples 1–4 above),
> `pev = TRUE` (the default) is always correct.

---

## Example 6: Adjusting `alpha`

A stricter significance level widens the criterion, reducing the number of
significant pairs.

```{r alpha_adjust}
#| cache: true
r_05 <- compare(model, term = "Variety:Site", by = "Site",
                type = "HSD", alpha = 0.05)
r_01 <- compare(model, term = "Variety:Site", by = "Site",
                type = "HSD", alpha = 0.01)

env1_05 <- unique(r_05$HSD[r_05$Site == "Env1"])
env1_01 <- unique(r_01$HSD[r_01$Site == "Env1"])
cat("Env1  alpha=0.05 HSD:", round(env1_05, 1L), "\n")
cat("Env1  alpha=0.01 HSD:", round(env1_01, 1L), "\n")
```

---

## Visualisation with `plot_compare()`

`plot_compare()` reads the `compare()` result directly and infers the grouping
structure automatically — no extra arguments are needed to handle `by` groups.

### Dot plot — `type = "dotplot"`

Varieties are sorted by predicted value, highest at the top.  A **shaded band**
of width equal to the HSD is drawn below the top-ranked variety.  Any variety
within the band is not significantly different from the best; those outside
(coloured red) are.  The dashed vertical line marks the top-ranked predicted
value.

```{r dotplot_single}
#| fig-height: 7
p1 <- plot_compare(res_hsd |> subset(Site == "Env1"), type = "dotplot") +
  ggtitle("Variety comparison — Env1 (HSD)")
print(p1)
```

When a `by`-grouped result is passed, each group becomes its own facet panel
with its own site-specific criterion band:

```{r dotplot_grouped}
#| fig-height: 7
p2 <- plot_compare(res_hsd, type = "dotplot") +
  ggtitle("Variety comparison by site (HSD)",
          subtitle = "Band width = site-specific HSD")
print(p2)
```

### Reference variety — `reference =`

Setting `reference` centres the band on a named variety rather than the best.
Points are coloured three ways: **green** (significantly better), **blue**
(not significantly different), **red** (significantly worse).  The reference
variety is shown as a filled diamond.

This is the natural way to ask *"which varieties are significantly better or
worse than our commercial check?"*

```{r dotplot_reference}
#| fig-height: 7
# Use Var10 as the commercial check reference
p3 <- plot_compare(res_hsd, type = "dotplot", reference = "Var10") +
  ggtitle("Variety comparison — reference: Var10",
          subtitle = "Green = sig. better | Blue = not sig. | Red = sig. worse")
print(p3)
```

### Half-criterion error bars — `type = "errbar"`

Each variety is shown as a point (its predicted value) with a horizontal line
segment spanning $\hat{\tau} \pm \frac{1}{2}\text{HSD}$.  The visual
significance rule is exact: **two varieties are significantly different if and
only if their bars do not overlap**.  This is equivalent to the criterion test
$|\hat{\tau}_i - \hat{\tau}_j| > \text{HSD}$.

The error bar display is particularly useful when interest lies in all pairwise
comparisons rather than just comparisons with the best variety — every
non-overlapping pair stands out directly, not just those that fall outside a
single band.

```{r errbar_single}
#| fig-height: 7
p_eb1 <- plot_compare(res_hsd |> subset(Site == "Env1"), type = "errbar") +
  ggtitle("Half-HSD error bars — Env1",
          subtitle = "Non-overlapping bars = significant difference")
print(p_eb1)
```

With a `by`-grouped result, each site gets its own panel.  Bar widths differ
between sites because each site has its own spatial error variance and therefore
its own HSD.

```{r errbar_grouped}
#| fig-height: 7
p_eb2 <- plot_compare(res_hsd, type = "errbar") +
  ggtitle("Half-HSD error bars by site",
          subtitle = "Bar width = site-specific HSD/2")
print(p_eb2)
```

The criterion type changes the bar width directly — LSD bars are shorter
(more discriminating) and Bonferroni bars are wider (more conservative).  This
makes the effect of the multiplicity correction immediately visible:

```{r errbar_lsd}
#| fig-height: 7
p_eb3 <- plot_compare(res_lsd |> subset(Site == "Env1"), type = "errbar") +
  ggtitle("Half-LSD error bars — Env1",
          subtitle = "LSD bars are shorter than HSD bars — more pairs declared significant")
print(p_eb3)
```

### Compact letter display — `type = "letters"`

The compact letter display (CLD) is the standard reporting format in plant
science.  Varieties sharing at least one letter are not significantly different.
Letters are assigned by the descending sweep algorithm using the HSD as the
significance threshold.

```{r letters_single}
#| fig-height: 7
p4 <- plot_compare(res_hsd |> subset(Site == "Env1"), type = "letters") +
  ggtitle("Compact letter display — Env1 (HSD)")
print(p4)
```

With `by`-groups, each site gets its own panel with independent letter
assignments:

```{r letters_grouped}
#| fig-height: 7
p5 <- plot_compare(res_hsd, type = "letters") +
  ggtitle("CLD — 20 varieties by site",
          subtitle = "Letters are independent within each site")
print(p5)
```

### Pairwise significance heatmap — `type = "heatmap"`

The heatmap shows all $\binom{20}{2} = 190$ pairwise absolute differences as
a colour-coded matrix.  A white cross (×) marks pairs that exceed the HSD
criterion.  For 20+ varieties this reveals the full discrimination pattern at a
glance — something no letter display can convey.

```{r heatmap_variety}
#| fig-height: 7
#| fig-width: 10
p6 <- plot_compare(res_hsd, type = "heatmap") +
  ggtitle("Variety pairwise significance — 4 sites",
          subtitle = "White × = significant at HSD level")
print(p6)
```

### Interactive dotplot — `interactive = TRUE`

Setting `interactive = TRUE` returns a `pc_interactive` object.  When printed,
it renders an interactive **plotly** widget in the RStudio Viewer pane
(not the Graphics pane).  You can still use `+` to add ggplot2 layers before
the conversion.

The dotplot has a special hover interaction: hovering over any variety
**moves the criterion band** to be centred on that variety's predicted value
and **recolours all points** relative to it (green = significantly better,
blue = not significant, red = significantly worse).  Unhovering restores the
original state.

```{r interactive}
#| eval: true
# Single-site subset keeps the demo focused on the hover interaction
as_plotly(plot_compare(res_hsd |> subset(Site == "Env1"),
                       type = "dotplot", interactive = TRUE))
```

> **Note:** `interactive = TRUE` requires the **plotly** package
> (`install.packages("plotly")`).  The widget is embedded directly in the
> HTML document — hover over any variety point to move the criterion band
> and recolour all points relative to it.
>
> **Multi-panel use:** pass the full `by`-grouped result to get one facet
> per group.  `as_plotly()` automatically adjusts subplot spacing; set
> `height` (pixels) if you need more room, e.g.
> `as_plotly(plot_compare(res_hsd, ...), height = 700)`.

### Extracting plot data — `return_data = TRUE`

All three plot types support `return_data = TRUE`, which returns the tidy
data frame used to build the plot:

```{r return_data}
df_dot <- plot_compare(res_hsd, type = "dotplot", return_data = TRUE)
head(df_dot[, c("Group", "Variety", "pred", "crit", "rank", "sig")])
```

```{r return_data_heat}
df_heat <- plot_compare(res_hsd |> subset(Site == "Env1"),
                        type   = "heatmap",
                        return_data = TRUE)
head(df_heat[, c("Var_x", "Var_y", "diff", "sig")])
```

The ggplot object returned by `plot_compare()` can also be extended with `+`:

```{r extend}
#| fig-height: 7
plot_compare(res_hsd, type = "letters") +
  ggplot2::ggtitle("Tukey HSD — 20 varieties across 4 sites",
                   subtitle = "alpha = 0.05 | Letters independent per site") +
  ggplot2::xlab("Predicted yield (kg/ha)")
```

---

## Summary

### `compare()`

| Task | Specification |
|------|--------------|
| HSD (Tukey) criterion | `type = "HSD"` (default) |
| LSD (unadjusted) criterion | `type = "LSD"` |
| Bonferroni criterion | `type = "Bonferroni"` |
| One global group | `by = NULL` (default) |
| Single factor by-group | `by = "Site"` |
| Multi-factor by-group | `by = c("Site", "Treatment")` or `by = "Site:Treatment"` |
| Stricter significance level | `alpha = 0.01` |
| Use posterior variance for BLUPs | `pev = FALSE` |
| Pass extra args to `predict()` | `...` (e.g. `present = ...`) |

### `plot_compare()`

| Task | Specification |
|------|--------------|
| Sorted dot plot with criterion band | `type = "dotplot"` (default) |
| Half-criterion error bars (non-overlap = significant) | `type = "errbar"` |
| Compact letter display | `type = "letters"` |
| Pairwise significance heatmap | `type = "heatmap"` |
| Band anchored to best variety | `reference = NULL` (default) |
| Band anchored to specific variety | `reference = "VarName"` |
| Interactive plotly widget | `interactive = TRUE` |
| Extend before plotly conversion | `result + ggtitle(...)` |
| Export tidy data frame | `return_data = TRUE` |

---

## References

Tukey, J.W. (1949). Comparing individual means in the analysis of variance.
*Biometrics*, **5**(2), 99–114.

Piepho, H.P. (2004). An algorithm for a letter-based representation of all
pairwise comparisons. *Journal of Computational and Graphical Statistics*,
**13**(2), 456–466.

