---
title: "Multiple Comparison Criteria for Predicted Values"
subtitle: "`compare()` and `plot_compare()`"
author: "biomAid package"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
library(asreml)
```

---

## Overview

`compare()` computes a pairwise comparison criterion — **HSD**, **LSD**, or
**Bonferroni-corrected LSD** — for predicted values obtained from an ASReml-R
V4 model.  The criterion is a single number: two predicted values in the same
group are declared significantly different if and only if their absolute
difference exceeds it.

Three comparison types are available:

| Type | Formula | Multiple-testing control | Best used when |
|------|---------|--------------------------|----------------|
| `"HSD"` (Tukey) | $\frac{\overline{\text{SED}}}{\sqrt{2}} \times q_{1-\alpha}(n_g, \nu)$ | Family-wise (exact for balanced designs) | Routine variety/treatment comparisons |
| `"LSD"` | $\overline{\text{SED}} \times t_{\alpha/2,\,\nu}$ | None | Exploratory or preplanned single comparisons |
| `"Bonferroni"` | $\overline{\text{SED}} \times t_{\alpha/(2m),\,\nu}$ | Family-wise (conservative) | Small number of specific comparisons |

The companion function `plot_compare()` turns a `compare()` result directly
into publication-ready visualisations: a sorted **dot plot** with a
criterion-width band, a **compact letter display** (CLD), or a
**pairwise significance heatmap**.

---

## Mathematical Framework

### Prediction error variance and the average SED

Let $\hat{\boldsymbol{\tau}}_g = (\hat{\tau}_{1g}, \ldots, \hat{\tau}_{n_g g})^\top$
be the vector of $n_g$ predicted values in group $g$, with prediction error
variance–covariance matrix

$$
\mathbf{V}_g = \operatorname{Var}(\hat{\boldsymbol{\tau}}_g - \boldsymbol{\tau}_g),
$$

returned by `predict(model, classify = term, vcov = TRUE)$vcov` in ASReml-R V4.

The standard error of a difference (SED) for the pair $(i, j)$ is

$$
\text{SED}_{ij} = \sqrt{V_{ii} + V_{jj} - 2V_{ij}}.
$$

`compare()` uses the **arithmetic mean SED** across all $m = \binom{n_g}{2}$
pairs as the representative spread of prediction error:

$$
\overline{\text{SED}}_g
  = \frac{1}{m}\sum_{i < j} \text{SED}_{ij}.
$$

Using the average SED (rather than a single representative pair) is robust when
PEVs vary across units — as they do in unbalanced designs and spatial models.

---

### Comparison criteria

Let $\nu = \texttt{model\$nedf}$ be the error degrees of freedom.

**Tukey HSD** — controls the family-wise error rate (FWER) exactly for
balanced designs by referring to the Studentised range distribution:

$$
\text{HSD}_g
  = \frac{\overline{\text{SED}}_g}{\sqrt{2}} \times
    q_{1-\alpha}(n_g, \nu),
$$

where $q_{1-\alpha}(n_g, \nu)$ is the upper $\alpha$ critical value of the
Studentised range with $n_g$ means and $\nu$ degrees of freedom.

**LSD** — no multiplicity correction; equivalent to a two-sided $t$-test at
level $\alpha$ for each pair:

$$
\text{LSD}_g = \overline{\text{SED}}_g \times t_{\alpha/2,\,\nu}.
$$

**Bonferroni** — applies the Bonferroni correction across all $m$ pairs,
controlling the FWER conservatively:

$$
\text{Bonferroni}_g
  = \overline{\text{SED}}_g \times t_{\alpha / (2m),\,\nu}.
$$

For all three types: $|\hat{\tau}_i - \hat{\tau}_j| > \text{criterion}$
declares pair $(i, j)$ significantly different.

---

### Posterior variance (`pev = FALSE`)

When `term` is in the **random** part of the model, the PEV
$\mathbf{V}_{\hat{\boldsymbol{\tau}}}$ underestimates the uncertainty of the
difference between BLUPs.  Setting `pev = FALSE` substitutes the **posterior
variance**

$$
\operatorname{Var}(\hat{\boldsymbol{\tau}} \mid \mathbf{y})
  = \mathbf{G} - \mathbf{V}_{\hat{\boldsymbol{\tau}}}
$$

which accounts for the shrinkage in BLUP estimation and gives a more
appropriate measure of uncertainty for genotypic BLUPs.

---

## Setup

### Simulated data and model

> **Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
> needed to read it; run the code chunks only if you have ASReml-R installed.

We use `simTrialData()` to generate a balanced multi-environment trial with
20 varieties, 4 sites, and 3 nitrogen treatments (T0, T1, T2). Variety is in
the random effects so that BLUPs can be demonstrated alongside BLUEs.

```{r simulate}
#| cache: true
out <- simTrialData(
  nvar          = 20L,
  nsite         = 4L,
  treatments    = c("T0", "T1", "T2"),
  treat_effects = c(0, 80, 200),
  nrep          = 3L,
  seed          = 42L,
  verbose       = FALSE
)
dat        <- out$data
dat$Row    <- factor(dat$Row)
dat$Column <- factor(dat$Column)
```

We fit a model with treatments as fixed effects and variety as a random effect
so that both BLUEs (treatments) and BLUPs (varieties) are available for
comparison:

```{r fit_model}
#| cache: true
#| results: hide
asreml.options(trace = FALSE)
model <- asreml(
  yield    ~ Treatment * Site,
  random   = ~ Variety + at(Site):Rep,
  residual = ~ dsum(~ ar1(Row):ar1(Column) | Site),
  data     = dat,
  maxit    = 30L
)
model <- update(model)
```

---

## Example 1: HSD — single global group

The simplest use: compare all treatment BLUEs across the trial in a single
group, with no `by` argument.  `compare()` calls `predict()` internally using
`term` as the `classify`.

```{r hsd_single}
res1 <- compare(
  model,
  term = "Treatment",
  type = "HSD"
)

res1
```

The `HSD` column is **constant within the group** — all predicted values use
the same criterion.  Two treatments are significantly different if
`|pred_i - pred_j| > HSD`.  The `avsed` column records the average SED on
which the criterion is based.

Applying the criterion manually:

```{r hsd_single_apply}
# Are T0 and T2 significantly different?
preds <- res1$predicted.value
names(preds) <- res1$Treatment
cat("T0 vs T2: |diff| =", round(abs(preds["T0"] - preds["T2"]), 1L),
    "  HSD =", round(res1$HSD[1L], 1L),
    "  Significant:", abs(preds["T0"] - preds["T2"]) > res1$HSD[1L], "\n")
```

---

## Example 2: HSD by site

Adding `by = "Site"` performs independent comparisons within each site, using
each site's own prediction error variance matrix and thus its own criterion.

```{r hsd_by_site}
#| cache: true
res2 <- compare(
  model,
  term = "Treatment:Site",
  by   = "Site",
  type = "HSD"
)

# HSD varies by site (different PEV per site from the spatial model)
unique(res2[, c("Site", "HSD", "avsed")])
```

The HSD differs between sites because the spatial AR1×AR1 residual structure
gives each site a different effective error variance. Sites with more efficient
spatial control (lower `avsed`) will have a smaller, more powerful criterion.

---

## Example 3: Multi-factor `by`

When variety BLUPs are predicted across all treatment × site combinations,
passing a character vector to `by` splits comparisons by every combination
of both factors.  The colon-string form `"Site:Treatment"` is equivalent.

```{r hsd_multi_by}
#| cache: true
# Variety BLUPs: compare varieties within each Site × Treatment combination
res3a <- compare(model, term = "Variety:Treatment:Site",
                 by = c("Site", "Treatment"), type = "HSD")
res3b <- compare(model, term = "Variety:Treatment:Site",
                 by = "Site:Treatment", type = "HSD")

# HSD per Site × Treatment group (first 6 rows)
unique(res3a[, c("Site", "Treatment", "HSD", "avsed")])[1:6, ]
```

The criterion varies across Site × Treatment combinations because each cell
has a different effective error variance — the spatial model and treatment
design make each sub-environment unique.

---

## Example 4: LSD and Bonferroni

The three criteria differ only in the multiplier applied to `avsed`.
LSD is the most liberal (no correction); Bonferroni is the most conservative;
HSD sits between them for most group sizes.

```{r three_criteria}
#| cache: true
res_hsd <- compare(model, term = "Treatment:Site", by = "Site", type = "HSD")
res_lsd <- compare(model, term = "Treatment:Site", by = "Site", type = "LSD")
res_bon <- compare(model, term = "Treatment:Site", by = "Site",
                   type = "Bonferroni")

# Criteria for Env1
env1 <- function(r, col) round(unique(r[[col]][r$Site == "Env1"]), 1L)
cat("Env1 criteria:\n")
cat("  LSD        =", env1(res_lsd, "LSD"),        "\n")
cat("  HSD        =", env1(res_hsd, "HSD"),        "\n")
cat("  Bonferroni =", env1(res_bon, "Bonferroni"), "\n")
```

For three treatments, $m = \binom{3}{2} = 3$ pairs.  Bonferroni divides
$\alpha$ by 6 (two-sided), LSD divides by 2 — the difference grows rapidly
with the number of treatments.

---

## Example 5: Variety BLUPs with `pev = FALSE`

When comparing variety BLUPs, the PEV understates uncertainty because it
ignores the shrinkage towards zero inherent in random effects.  Setting
`pev = FALSE` substitutes the posterior variance $\mathbf{G} - \mathbf{PEV}$,
giving a more appropriate — and larger — criterion.

```{r pev_false}
#| cache: true
res_pev  <- compare(model, term = "Variety", type = "HSD", pev = TRUE)
res_post <- compare(model, term = "Variety", type = "HSD", pev = FALSE)

cat("HSD (PEV):            ", round(res_pev$HSD[1L],  1L), "\n")
cat("HSD (posterior var.): ", round(res_post$HSD[1L], 1L), "\n")
```

The posterior-variance criterion is larger, reflecting the true uncertainty
in declaring two BLUPs significantly different.

---

## Example 6: Adjusting `alpha`

A stricter significance level widens the criterion, reducing the number of
significant pairs.

```{r alpha_adjust}
#| cache: true
r_05 <- compare(model, term = "Treatment:Site", by = "Site",
                type = "HSD", alpha = 0.05)
r_01 <- compare(model, term = "Treatment:Site", by = "Site",
                type = "HSD", alpha = 0.01)

env1_05 <- unique(r_05$HSD[r_05$Site == "Env1"])
env1_01 <- unique(r_01$HSD[r_01$Site == "Env1"])
cat("Env1  alpha=0.05 HSD:", round(env1_05, 1L), "\n")
cat("Env1  alpha=0.01 HSD:", round(env1_01, 1L), "\n")
```

---

## Visualisation with `plot_compare()`

`plot_compare()` reads the `compare()` result directly and infers the grouping
structure automatically — no extra arguments are needed to handle `by` groups.

### Dot plot — `type = "dotplot"`

Varieties (or treatments) are sorted by predicted value, highest at the top.
A **shaded band** of width equal to the criterion is drawn below the
top-ranked variety.  Any variety within the band is not significantly different
from the best; those outside (coloured red) are.  The dashed vertical line
marks the top-ranked predicted value.

```{r dotplot_single}
#| fig-height: 5
p1 <- plot_compare(res_hsd |> subset(Site == "Env1"), type = "dotplot") +
  ggtitle("Treatment comparison — Env1 (HSD)")
print(p1)
```

When a `by`-grouped result is passed, each group becomes its own facet panel
with its own site-specific criterion band:

```{r dotplot_grouped}
#| fig-height: 5
p2 <- plot_compare(res_hsd, type = "dotplot") +
  ggtitle("Treatment comparison by site (HSD)",
          subtitle = "Band width = site-specific HSD")
print(p2)
```

### Reference variety — `reference =`

Setting `reference` centres the band on a specific variety rather than the
best.  Points are coloured three ways: **green** (significantly better than
the reference), **blue** (not significantly different), **red** (significantly
worse).  The reference variety is shown as a filled diamond.

This is the natural way to ask *"which varieties are significantly better or
worse than our commercial check?"*

```{r dotplot_reference}
#| fig-height: 6
# Use variety BLUPs for a more meaningful comparison
res_var <- compare(model, term = "Variety:Site", by = "Site",
                   type = "HSD")

# Pick the median-ranked variety in Env1 as the reference check
env1_preds <- res_var[res_var$Site == "Env1", ]
ref_var    <- env1_preds$Variety[order(env1_preds$predicted.value)[10L]]
cat("Reference variety:", as.character(ref_var), "\n")

p3 <- plot_compare(res_var, type = "dotplot", reference = as.character(ref_var)) +
  ggtitle(paste0("Variety comparison — reference: ", ref_var),
          subtitle = "Green = sig. better | Blue = not sig. | Red = sig. worse")
print(p3)
```

### Compact letter display — `type = "letters"`

The compact letter display (CLD) is the standard reporting format in plant
science.  Varieties sharing at least one letter are not significantly different.
Letters are assigned by the descending sweep algorithm using the comparison
criterion as the significance threshold.

```{r letters_single}
#| fig-height: 5
p4 <- plot_compare(res_hsd |> subset(Site == "Env1"), type = "letters") +
  ggtitle("Compact letter display — Env1 (HSD)")
print(p4)
```

With `by`-groups, each site gets its own panel with independent letter
assignments:

```{r letters_grouped}
#| fig-height: 5
p5 <- plot_compare(res_hsd, type = "letters") +
  ggtitle("CLD by site", subtitle = "Letters are independent within each site")
print(p5)
```

### Pairwise significance heatmap — `type = "heatmap"`

The heatmap shows all $\binom{n}{2}$ pairwise absolute differences as a
colour-coded matrix.  A white cross (×) marks pairs that exceed the criterion
(significant).  For large variety sets this reveals the full pattern of
discrimination in a single view — something no letter display can show.

```{r heatmap_single}
#| fig-height: 5
p6 <- plot_compare(res_hsd |> subset(Site == "Env1"), type = "heatmap") +
  ggtitle("Pairwise significance heatmap — Env1")
print(p6)
```

```{r heatmap_grouped}
#| fig-height: 5
#| fig-width: 9
p7 <- plot_compare(res_hsd, type = "heatmap") +
  ggtitle("Pairwise significance heatmap by site")
print(p7)
```

### Variety heatmap — larger set

The heatmap is especially useful when there are many varieties and the letter
display becomes long.

```{r heatmap_variety}
#| fig-height: 7
#| fig-width: 9
p8 <- plot_compare(res_var, type = "heatmap") +
  ggtitle("Variety pairwise significance — all sites",
          subtitle = "White × = significant at HSD level")
print(p8)
```

### Interactive dotplot — `interactive = TRUE`

Setting `interactive = TRUE` returns a `pc_interactive` object.  When printed,
it renders an interactive **plotly** widget in the RStudio Viewer pane
(not the Graphics pane).  You can still use `+` to add ggplot2 layers before
the conversion.

The dotplot has a special hover interaction: hovering over any variety
**moves the criterion band** to be centred on that variety's predicted value
and **recolours all points** relative to it (green = significantly better,
blue = not significant, red = significantly worse).  Unhovering restores the
original state.

```{r interactive}
#| eval: false
# Renders to the RStudio Viewer pane — interactive hover moves the band
plot_compare(res_var, type = "dotplot", interactive = TRUE) +
  ggplot2::ggtitle("Hover over any variety to explore significance")

# Also works for letters and heatmap (tooltips, no band interaction)
plot_compare(res_hsd, type = "heatmap", interactive = TRUE)
```

> **Note:** `interactive = TRUE` requires the **plotly** package
> (`install.packages("plotly")`).  Interactive plots render to the Viewer pane
> as an HTML widget, not to the standard R Graphics window.

### Extracting plot data — `return_data = TRUE`

All three plot types support `return_data = TRUE`, which returns the tidy
data frame used to build the plot:

```{r return_data}
df_dot <- plot_compare(res_hsd, type = "dotplot", return_data = TRUE)
head(df_dot[, c("Group", "Variety", "pred", "crit", "rank", "sig")])
```

```{r return_data_heat}
df_heat <- plot_compare(res_hsd |> subset(Site == "Env1"),
                        type = "heatmap", return_data = TRUE)
head(df_heat[, c("Var_x", "Var_y", "diff", "sig")])
```

The data frame can be passed directly into any `ggplot2` workflow:

```{r bespoke}
#| fig-height: 5
df_env1 <- plot_compare(res_hsd |> subset(Site == "Env1"),
                        type = "dotplot", return_data = TRUE)

ggplot(df_env1, aes(x = pred, y = reorder(Variety, pred),
                     colour = sig)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = band_lo, xmax = band_hi),
                 height = 0, linewidth = 0.4, colour = "grey60") +
  scale_colour_manual(values = c(ns = "#4E79A7", sig = "#E15759"),
                      name = NULL) +
  labs(x = "Predicted value", y = NULL,
       title = "Custom: dot + criterion bar plot (Env1)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The ggplot object returned by `plot_compare()` can also be extended with `+`:

```{r extend}
#| fig-height: 5
plot_compare(res_hsd, type = "letters") +
  ggplot2::ggtitle("Tukey HSD letter groups by site",
                   subtitle = paste0("alpha = 0.05 | ",
                                     nrow(out$data) / 4L,
                                     " plots per site")) +
  ggplot2::xlab("Predicted yield (kg/ha)")
```

---

## Summary

### `compare()`

| Task | Specification |
|------|--------------|
| HSD (Tukey) criterion | `type = "HSD"` (default) |
| LSD (unadjusted) criterion | `type = "LSD"` |
| Bonferroni criterion | `type = "Bonferroni"` |
| One global group | `by = NULL` (default) |
| Single factor by-group | `by = "Site"` |
| Multi-factor by-group | `by = c("Site", "Treatment")` or `by = "Site:Treatment"` |
| Stricter significance level | `alpha = 0.01` |
| Use posterior variance for BLUPs | `pev = FALSE` |
| Pass extra args to `predict()` | `...` (e.g. `present = ...`) |

### `plot_compare()`

| Task | Specification |
|------|--------------|
| Sorted dot plot with criterion band | `type = "dotplot"` (default) |
| Compact letter display | `type = "letters"` |
| Pairwise significance heatmap | `type = "heatmap"` |
| Band anchored to best variety | `reference = NULL` (default) |
| Band anchored to specific variety | `reference = "VarName"` |
| Interactive plotly widget | `interactive = TRUE` |
| Extend before plotly conversion | `result + ggtitle(...)` |
| Export tidy data frame | `return_data = TRUE` |

---

## References

Tukey, J.W. (1949). Comparing individual means in the analysis of variance.
*Biometrics*, **5**(2), 99–114.

Piepho, H.P. (2004). An algorithm for a letter-based representation of all
pairwise comparisons. *Journal of Computational and Graphical Statistics*,
**13**(2), 456–466.

Gilmour, A.R., Gogel, B.J., Cullis, B.R. & Thompson, R. (2009).
*ASReml User Guide Release 3.0*. VSN International Ltd, Hemel Hempstead, UK.
