---
title: "Factor Analytic Selection Tools: FAST and iClass"
subtitle: "`fastIC()` and `plot_fastIC()`"
author: "Julian Taylor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
asreml_ok <- requireNamespace("asreml",    quietly = TRUE) &&
             requireNamespace("ASExtras4", quietly = TRUE)
if (asreml_ok) {
  library(asreml)
  library(ASExtras4)
  asreml.options(trace = FALSE)
}
```

---

## Overview

`fastIC()` provides a unified implementation of two complementary approaches
for summarising variety performance from a **Factor Analytic (FA) linear
mixed model** fitted in ASReml-R V4:

- **FAST** (Factor Analytic Selection Tools; Smith & Cullis, 2018) — global
  Overall Performance and stability across all environments.
- **iClass** (interaction class; Smith et al., 2021) — within-class
  performance and stability for environments grouped by their loading
  sign patterns.

Both frameworks are treated as a single continuum. FAST global metrics
(`global_op`, `global_stab`) are always computed across **all** environments.
iClass metrics (`iclass`, `iClassOP`, `iClassRMSD`) partition environments
by the sign pattern of their first `ic.num` rotated loadings. When
`ic.num = 1` and — as is typical after rotation — all first-factor
loadings are positive, there is exactly one iClass and the two frameworks
agree exactly: `iClassOP = global_op` and `iClassRMSD = global_stab`.

`plot_fastIC()` provides seven complementary visualisations:

| Type | Description |
|------|-------------|
| `"fast"` | Global OP vs stability scatter with quadrant annotations |
| `"biplot"` | FA biplot: genotype scores + environment loading arrows |
| `"CVE"` | Common Variety Effect heatmap (genotype × environment) |
| `"dev"` | Residual deviation heatmap (GEI beyond Factor 1) |
| `"iclass"` | iClassOP vs iClassRMSD scatter, faceted by interaction class |
| `"OP.pairs"` | Lower-triangular pairs plot of iClassOP across all classes |
| `"OP.variety"` | iClassOP trajectory for highlighted varieties across classes |

---

## Mathematical Framework

### Factor Analytic mixed model

Let $\hat{\mathbf{U}} \in \mathbb{R}^{n \times t}$ be the matrix of variety
BLUPs across $n$ varieties and $t$ environments, obtained from an FA($k$)
random structure fitted to the Site × Variety term. The FA model
decomposes the genetic covariance as
$\mathbf{G} = \hat{\boldsymbol{\Lambda}}\hat{\boldsymbol{\Lambda}}^\top +
\hat{\boldsymbol{\Psi}}$, where
$\hat{\boldsymbol{\Psi}} = \operatorname{diag}(\hat{\psi}_1,\ldots,\hat{\psi}_t)$
is the diagonal matrix of specific variances.

After varimax rotation, `ASExtras4::fa.asreml()` returns:

- **Rotated loadings** $\hat{\boldsymbol{\Lambda}} \in \mathbb{R}^{t \times k}$
  — element $\hat{\lambda}_{rj}$ is the loading of factor $r$ on environment
  $j$. After rotation, factor 1 is the general (non-crossover) factor with
  typically all-positive loadings; higher factors are bipolar.
- **Score EBLUPs** $\hat{\mathbf{F}} \in \mathbb{R}^{n \times k}$ — element
  $\hat{f}_{rg}$ is the EBLUP of variety $g$ on factor $r$. The score
  vector for variety $g$ is $\hat{\mathbf{f}}_g = (\hat{f}_{1g}, \ldots,
  \hat{f}_{kg})^\top$.
- **Specific variances** $\hat{\psi}_j$ — the environment-specific residual
  genetic variance for environment $j$.

The **Common Variety Effect** (CVE) and **Variety Effect** (VE) for variety
$g$ in environment $j$ are:

$$
\widehat{\text{CVE}}(g,j)
  = \hat{\boldsymbol{\lambda}}_j^\top \hat{\mathbf{f}}_g
  = \sum_{r=1}^{k} \hat{\lambda}_{rj}\,\hat{f}_{rg},
\qquad
\widehat{\text{VE}}(g,j) = \widehat{\text{CVE}}(g,j) + \hat{\psi}_j,
$$

where $\hat{\boldsymbol{\lambda}}_j = (\hat{\lambda}_{1j}, \ldots,
\hat{\lambda}_{kj})^\top$ is the loading vector for environment $j$.
In matrix form, the full $n \times t$ CVE matrix is
$\widehat{\mathbf{CVE}} = \hat{\mathbf{F}}\hat{\boldsymbol{\Lambda}}^\top$.

### Global FAST metrics

The global metrics summarise each variety across **all** $t$ environments
and are invariant to the iClass classification. Define
$\bar{\boldsymbol{\lambda}}_r = t^{-1}\mathbf{1}_t^\top\hat{\boldsymbol{\Lambda}}_{*r}$
as the mean of the $r$th column of $\hat{\boldsymbol{\Lambda}}$. Then:

$$
\text{global\_op}(g)
  = \bar{\lambda}_{1} \cdot \hat{f}_{1g}
  = t^{-1}\bigl(\mathbf{1}_t^\top\hat{\boldsymbol{\lambda}}_{\cdot 1}\bigr)\hat{f}_{1g},
$$

which is on the same scale as the trait and represents expected performance
at an average environment. The **global deviation** vector and **global
stability** for variety $g$ are (for $k > 1$):

$$
\mathbf{d}_g = \widehat{\mathbf{CVE}}_{g\cdot} - \hat{f}_{1g}\,\hat{\boldsymbol{\lambda}}_{\cdot 1}^\top
  \in \mathbb{R}^t,
\qquad
\text{global\_stab}(g) = \|\mathbf{d}_g\|_2 / \sqrt{t}
  = \sqrt{t^{-1}\,\mathbf{d}_g^\top\mathbf{d}_g},
$$

where $\widehat{\mathbf{CVE}}_{g\cdot}$ is the $g$th row of the CVE matrix
and $\hat{\boldsymbol{\lambda}}_{\cdot 1}$ is the first column of
$\hat{\boldsymbol{\Lambda}}$. A small `global_stab` indicates broad
adaptation; larger values signal environment-specific response (crossover GEI).

### iClass metrics

Environments are grouped by the sign pattern of their first `ic.num`
rotated loadings. Let
$\hat{\boldsymbol{\Lambda}}_{[q]} \in \mathbb{R}^{t \times q}$,
$q = \texttt{ic.num}$, denote the first $q$ columns of
$\hat{\boldsymbol{\Lambda}}$. The iClass label for environment $j$ is:

$$
\text{iClass}(j) = \bigotimes_{r=1}^{q}
\begin{cases} \textup{p} & \hat{\lambda}_{rj} \ge 0 \\ \textup{n} & \hat{\lambda}_{rj} < 0 \end{cases}.
$$

For each iClass $\omega$ let $\mathbf{E}_\omega$ denote the index set of
environments in $\omega$ and define the **class mean loading vector**
$\bar{\boldsymbol{\lambda}}_\omega = |\omega|^{-1}
\sum_{j\in\mathbf{E}_\omega} \hat{\boldsymbol{\lambda}}_{j,[q]}
\in \mathbb{R}^q$.
The **iClassOP** for variety $g$ in class $\omega$ is:

$$
\text{iClassOP}(g,\omega)
  = \bar{\boldsymbol{\lambda}}_\omega^\top \hat{\mathbf{f}}_{g,[q]}
  = \sum_{r=1}^{q} \bar{\lambda}_{r\omega}\,\hat{f}_{rg},
$$

where $\hat{\mathbf{f}}_{g,[q]} = (\hat{f}_{1g},\ldots,\hat{f}_{qg})^\top$
are the first $q$ scores for variety $g$. The **iClassRMSD** uses the
remaining factors as a residual:

$$
\text{iClassRMSD}(g,\omega)
  = \frac{1}{\sqrt{|\omega|}}\,
    \bigl\|\widehat{\mathbf{CVE}}_{g,\mathbf{E}_\omega}
    - \hat{\mathbf{F}}_{g,[q]}\hat{\boldsymbol{\Lambda}}_{\mathbf{E}_\omega,[q]}^\top
    \bigr\|_2,
$$

where the residual $\mathbf{dev}_{ic}(g,\omega) = \widehat{\mathbf{CVE}}_{g,\mathbf{E}_\omega}
- \hat{\mathbf{F}}_{g,[q]}\hat{\boldsymbol{\Lambda}}_{\mathbf{E}_\omega,[q]}^\top$
is the within-class GEI unexplained by the first $q$ factors. The kth
factor is **reserved for this residual**; therefore
$1 \le \texttt{ic.num} \le k - 1$.

### Relationship between FAST and iClass

When $\texttt{ic.num} = 1$ and all first-factor loadings are positive (one
iClass containing all environments):

$$
\text{iClassOP}(g,\{\text{all envs}\}) = \text{global\_op}(g),
\qquad
\text{iClassRMSD}(g,\{\text{all envs}\}) = \text{global\_stab}(g).
$$

When $\texttt{ic.num} \ge 2$ environments split into multiple classes.
`iClassOP` then captures a variety's expected performance **within
environments of each class**, isolating crossover GEI that `global_op`
averages away.

---

## Example 1 — Balanced MET: 50 Varieties × 12 Sites, FA(3), ic.num = 2

### Data simulation

> **Requires:** ASReml-R V4 and ASExtras4. The vignette is pre-rendered so
> no licence is needed to read it; run the code chunks only if you have the
> packages installed.

We simulate a balanced MET with pronounced Site × Variety interaction by
using a narrow correlation range for the genetic covariance matrix.
With `g_cor_min = 0.10` and `g_cor_max = 0.65`, genetic correlations span a
wide range, generating both positive-positive and positive-negative loading
patterns on higher FA factors — the precondition for meaningful iClass
structure.

```{r sim1}
#| cache: true
#| eval: !expr asreml_ok
set.seed(101L)
sim1 <- simTrialData(
  nvar        = 50L,
  nsite       = 12L,
  nrep        = 3L,
  G           = "auto",
  incidence   = "balanced",
  seed        = 101L,
  verbose     = FALSE,
  sim.options = list(
    g_cor_min     = 0.10,
    g_cor_max     = 0.65,
    sigma_genetic = 300,
    error_sd      = 280,
    site_mean     = 4500,
    site_sd       = 450
  )
)
dat1         <- sim1$data
dat1$Site    <- factor(dat1$Site)
dat1$Variety <- factor(dat1$Variety)
dat1$Rep     <- factor(dat1$Rep)
dat1$Row     <- factor(dat1$Row)
dat1$Column  <- factor(dat1$Column)
```

A quick look at the true genetic correlation structure confirms meaningful
variability — some environments are quite poorly correlated with others,
providing scope for crossover interaction.

```{r plot_sim1_cor}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_simTrialData(sim1, type = "correlation")
```

### Fitting FA(1), FA(2) and FA(3) models

We fit three models of increasing complexity, using `at(Site):Rep` for
replicate effects within sites:

```{r fit1_fa1}
#| cache: true
#| dependson: sim1
#| eval: !expr asreml_ok
#| results: hide
m1_fa1 <- asreml(
  yield  ~ Site,
  random = ~ fa(Site, 1):Variety + at(Site):Rep,
  data   = dat1,
  maxit  = 30L
)
m1_fa1 <- update(m1_fa1)
```

```{r fit1_fa2}
#| cache: true
#| dependson: fit1_fa1
#| eval: !expr asreml_ok
#| results: hide
m1_fa2 <- asreml(
  yield  ~ Site,
  random = ~ fa(Site, 2):Variety + at(Site):Rep,
  data   = dat1,
  maxit  = 30L
)
m1_fa2 <- update(m1_fa2)
```

```{r fit1_fa3}
#| cache: true
#| dependson: fit1_fa2
#| eval: !expr asreml_ok
#| results: hide
m1_fa3 <- asreml(
  yield  ~ Site,
  random = ~ fa(Site, 3):Variety + at(Site):Rep,
  data   = dat1,
  maxit  = 30L
)
m1_fa3 <- update(m1_fa3)
```

AIC confirms that the FA(3) model provides the best fit:

```{r aic1}
#| eval: !expr asreml_ok
.aic <- function(m) -2 * m$loglik + 2 * nrow(summary(m)$varcomp)
data.frame(
  Model  = c("FA(1)", "FA(2)", "FA(3)"),
  LogLik = round(c(m1_fa1$loglik, m1_fa2$loglik, m1_fa3$loglik), 2),
  npar   = c(nrow(summary(m1_fa1)$varcomp),
             nrow(summary(m1_fa2)$varcomp),
             nrow(summary(m1_fa3)$varcomp)),
  AIC    = round(c(.aic(m1_fa1), .aic(m1_fa2), .aic(m1_fa3)), 2)
)
```

### Running `fastIC()`

We run `fastIC()` on the FA(3) model with `ic.num = 2`. The first two
rotated factors define the iClass labels — environments with loading sign
pattern `"pp"` (positive on both factors) form one class, those with
pattern `"pn"` form another. Factor 3 is reserved for iClassRMSD.

```{r fastIC1}
#| cache: true
#| dependson: fit1_fa3
#| eval: !expr asreml_ok
res1 <- fastIC(
  model  = m1_fa3,
  term   = "fa(Site, 3):Variety",
  ic.num = 2L
)
```

```{r show_res1}
#| eval: !expr asreml_ok
# iClass assignments for each site
unique(res1[, c("Site", "iclass")])[order(unique(res1[, c("Site","iclass")])$iclass), ]
```

```{r iclass_sizes1}
#| eval: !expr asreml_ok
cat("iClass sizes (number of environments):\n")
print(table(unique(res1[, c("Site","iclass")])$iclass))
```

The two iClasses partition sites by whether they respond positively or
negatively to the second FA factor — a measure of crossover GEI that is
invisible from the global rankings alone.

### FAST: Global OP vs Stability

The FAST plot provides the traditional entry point. Varieties in the
upper-right quadrant are high-performing **and** broadly adapted; those in
the upper-left combine good average performance with more environment-specific
response.

```{r plot1_fast}
#| eval: !expr asreml_ok
#| fig-height: 5.5
plot_fastIC(res1, type = "fast", n_highlight = 5L)
```

### FA biplot

The biplot places genotypes in the space of factor scores (1 vs 2) and
environments as arrows scaled by their loadings. Environments coloured by
iClass visually confirm the separation: `"pp"` sites point into the positive
upper-right; `"pn"` sites point into the positive-negative quadrant.
Varieties closest to an environment arrow tend to perform well in that
environment.

```{r plot1_biplot}
#| eval: !expr asreml_ok
#| fig-height: 6
plot_fastIC(res1, type = "biplot", n_highlight = 5L)
```

### CVE heatmap

The Common Variety Effect heatmap shows the full genotype × environment
performance surface. Environments are grouped by iClass (column facets) and
ordered by first-factor loading within each class, while varieties are ordered
by `global_op` from top (best) to bottom (poorest). The diverging palette
makes crossover patterns immediately visible — varieties that are strongly
positive in one class but near-zero or negative in the other.

```{r plot1_cve}
#| eval: !expr asreml_ok
#| fig-height: 7
plot_fastIC(res1, type = "CVE")
```

### Residual deviation heatmap

The `"dev"` heatmap shows `global_dev` — the residual CVE after removing the
first-factor contribution. Cells with large absolute values indicate
varieties whose performance deviates most from the global trend in specific
environments.

```{r plot1_dev}
#| eval: !expr asreml_ok
#| fig-height: 7
plot_fastIC(res1, type = "dev")
```

### iClass: Within-class OP vs RMSD

Faceting by iClass separates the OP vs RMSD trade-off within each
interaction class. The vertical dashed line marks the mean iClassOP within
the class. Varieties above-right of the intersection are class champions —
high within-class performance with low within-class instability.

```{r plot1_iclass}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_fastIC(res1, type = "iclass", n_highlight = 4L)
```

### iClassOP pairs plot

With two iClasses the `"OP.pairs"` plot is a single scatter of iClassOP in
class `"pn"` (x-axis) against class `"pp"` (y-axis). The dashed diagonal
marks equal performance across classes. Varieties well above the line are
specialists for `"pp"` environments; those below favour `"pn"` environments.
Points near the diagonal are generalists.

```{r plot1_pairs}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_fastIC(res1, type = "OP.pairs", n_highlight = 5L)
```

### Variety trajectory

The `"OP.variety"` plot traces each variety's iClassOP across the ordered
interaction classes. Highlighted varieties are selected by default as those
with the largest range in iClassOP (i.e. the greatest crossover). Steep
crossing lines indicate strong specialists; near-parallel lines indicate
stable generalists.

```{r plot1_trajectory}
#| eval: !expr asreml_ok
#| fig-height: 5
plot_fastIC(res1, type = "OP.variety", n_highlight = 6L)
```

---

## Example 2 — Unbalanced MET: 75 Varieties × 25 Sites, FA(4), ic.num = 3

When the number of environments is larger and the genetic covariance
structure is more complex, higher-order FA models and more interaction
classes are needed to fully characterise GEI. Here we use an unbalanced
incidence structure to reflect a realistic breeding programme where not all
varieties are evaluated at every site.

### Data simulation

```{r sim2}
#| cache: true
#| eval: !expr asreml_ok
set.seed(202L)
sim2 <- simTrialData(
  nvar        = 75L,
  nsite       = 25L,
  nrep        = 2L,
  G           = "auto",
  incidence   = "unbalanced",
  seed        = 202L,
  verbose     = FALSE,
  sim.options = list(
    g_cor_min     = 0.05,
    g_cor_max     = 0.60,
    sigma_genetic = 320,
    error_sd      = 300,
    site_mean     = 4500,
    site_sd       = 500,
    core_pct      = 0.25,
    core_min_sites_pct = 0.80,
    reg_min_sites_pct  = 0.35,
    reg_max_sites_pct  = 0.75
  )
)
dat2         <- sim2$data
dat2$Site    <- factor(dat2$Site)
dat2$Variety <- factor(dat2$Variety)
dat2$Rep     <- factor(dat2$Rep)
dat2$Row     <- factor(dat2$Row)
dat2$Column  <- factor(dat2$Column)
```

The unbalanced incidence creates a core set of varieties present across
most sites and a peripheral set with sparser coverage:

```{r plot_sim2_incidence}
#| eval: !expr asreml_ok
#| fig-height: 7
plot_simTrialData(sim2, type = "incidence")
```

Connectivity summary:

```{r connectivity2}
#| eval: !expr asreml_ok
inc <- sim2$params$incidence
cat("Varieties per site — mean:", round(mean(colSums(inc)), 1),
    " range [", min(colSums(inc)), ",", max(colSums(inc)), "]\n")
cat("Sites per variety  — mean:", round(mean(rowSums(inc)), 1),
    " range [", min(rowSums(inc)), ",", max(rowSums(inc)), "]\n")
```

The true genetic correlation structure shows considerable variability
across the 25 sites, with some site pairs sharing very low or even
negative genetic correlations — precisely the conditions under which
four FA factors and three iClasses are warranted.

```{r plot_sim2_cor}
#| eval: !expr asreml_ok
#| fig-height: 5.5
plot_simTrialData(sim2, type = "correlation")
```

### Fitting FA(2), FA(3) and FA(4) models

```{r fit2_fa2}
#| cache: true
#| dependson: sim2
#| eval: !expr asreml_ok
#| results: hide
m2_fa2 <- asreml(
  yield  ~ Site,
  random = ~ fa(Site, 2):Variety + at(Site):Rep,
  data   = dat2,
  maxit  = 30L
)
m2_fa2 <- update(m2_fa2)
```

```{r fit2_fa3}
#| cache: true
#| dependson: fit2_fa2
#| eval: !expr asreml_ok
#| results: hide
m2_fa3 <- asreml(
  yield  ~ Site,
  random = ~ fa(Site, 3):Variety + at(Site):Rep,
  data   = dat2,
  maxit  = 30L
)
m2_fa3 <- update(m2_fa3)
```

```{r fit2_fa4}
#| cache: true
#| dependson: fit2_fa3
#| eval: !expr asreml_ok
#| results: hide
m2_fa4 <- asreml(
  yield  ~ Site,
  random = ~ fa(Site, 4):Variety + at(Site):Rep,
  data   = dat2,
  maxit  = 30L
)
m2_fa4 <- update(m2_fa4)
```

```{r aic2}
#| eval: !expr asreml_ok
data.frame(
  Model  = c("FA(2)", "FA(3)", "FA(4)"),
  LogLik = round(c(m2_fa2$loglik, m2_fa3$loglik, m2_fa4$loglik), 2),
  npar   = c(nrow(summary(m2_fa2)$varcomp),
             nrow(summary(m2_fa3)$varcomp),
             nrow(summary(m2_fa4)$varcomp)),
  AIC    = round(c(.aic(m2_fa2), .aic(m2_fa3), .aic(m2_fa4)), 2)
)
```

### Running `fastIC()` with ic.num = 3

With a FA(4) model and `ic.num = 3`, the first three factors define up to
$2^3 = 8$ potential iClasses, though only those sign patterns that are
actually represented by at least one environment will appear. Factor 4 is
reserved for iClassRMSD.

```{r fastIC2}
#| cache: true
#| dependson: fit2_fa4
#| eval: !expr asreml_ok
res2 <- fastIC(
  model  = m2_fa4,
  term   = "fa(Site, 4):Variety",
  ic.num = 3L
)
```

```{r show_iclasses2}
#| eval: !expr asreml_ok
# iClass assignments
ic_tab <- unique(res2[, c("Site", "iclass")])
ic_tab <- ic_tab[order(ic_tab$iclass, ic_tab$Site), ]
cat("iClass sizes (number of environments):\n")
print(table(ic_tab$iclass))
```

With three classification factors and 25 environments, several distinct
interaction classes emerge. Classes reflect groups of environments that share
the same pattern of positive/negative response on all three bipolar factors.

### FAST: Global OP vs Stability

```{r plot2_fast}
#| eval: !expr asreml_ok
#| fig-height: 5.5
plot_fastIC(res2, type = "fast", n_highlight = 6L)
```

### FA biplot

The biplot for the larger 25-site dataset shows a richer loading structure,
with environments spread more widely through the (factor 1, factor 2)
loading space and clearly distinguished by iClass colour.

```{r plot2_biplot}
#| eval: !expr asreml_ok
#| fig-height: 6.5
plot_fastIC(res2, type = "biplot", n_highlight = 5L)
```

### CVE heatmap

With four iClasses the CVE heatmap is faceted into four columns, each
grouping the sites that share the same three-factor loading sign pattern.
The within-class rank ordering of varieties often differs strikingly from
the global ranking, which motivates the iClass approach.

```{r plot2_cve}
#| eval: !expr asreml_ok
#| fig-height: 8
plot_fastIC(res2, type = "CVE")
```

### iClass OP vs RMSD

Faceted across all iClasses, this plot highlights that the within-class
RMSD for most varieties is considerably smaller than `global_stab` —
evidence that the crossover GEI captured by the first three factors has
been successfully isolated into the class structure.

```{r plot2_iclass}
#| eval: !expr asreml_ok
#| fig-height: 6
plot_fastIC(res2, type = "iclass", n_highlight = 4L)
```

### Lower-triangular iClassOP pairs plot

With four iClasses the `"OP.pairs"` plot becomes a $3 \times 3$
lower-triangular matrix of pairwise iClassOP scatters — one panel for each
combination of classes. Each panel uses the same diagonal reference line;
points far from the diagonal are class specialists, while points near it
perform similarly across both classes. The iClass structure is most
informative when the cloud of points in each panel shows clear deviation
from the diagonal in a consistent direction.

```{r plot2_pairs}
#| eval: !expr asreml_ok
#| fig-width: 10
#| fig-height: 9
plot_fastIC(res2, type = "OP.pairs", n_highlight = 5L)
```

### Variety trajectory

The trajectory plot is particularly striking with multiple iClasses. Each
line traces a variety's iClassOP from the first class (left) through to the
last (right). Strongly crossing lines — and especially varieties that switch
rank — illustrate precisely the crossover GEI that a global OP ranking
misses. The highlighted varieties are automatically chosen as those with the
greatest range in iClassOP.

```{r plot2_trajectory}
#| eval: !expr asreml_ok
#| fig-height: 5.5
plot_fastIC(res2, type = "OP.variety", n_highlight = 8L)
```

---

## Summary

| Feature | Example 1 | Example 2 |
|---------|-----------|-----------|
| Varieties | 50 | 75 |
| Sites | 12 | 25 |
| Incidence | Balanced | Unbalanced |
| FA order | FA(3) | FA(4) |
| `ic.num` | 2 | 3 |
| iClasses | 2 | up to 4 |
| Key plots | FAST, biplot, CVE, pairs | All 7 types |

The two examples illustrate a progression from a moderate-GEI balanced trial
where two iClasses cleanly separate environments, to a larger unbalanced
programme where three classification factors and up to four iClasses reveal a
richer GEI landscape. In both cases, the `"OP.pairs"` and `"OP.variety"` plots
are the most direct tools for identifying class specialists — varieties whose
ranking changes substantially depending on the interaction class of the
target environment.

---

## References

Smith, A.B. & Cullis, B.R. (2018). Plant breeding selection tools built on
factor analytic mixed models for multi-environment trial data.
*Euphytica*, **214**, 143.

Smith, A.B., Norman, A., Kuchel, H. & Cullis, B.R. (2021). Plant variety
selection using interaction classes derived from factor analytic linear
mixed models: models with independent variety effects.
*Frontiers in Plant Science*, **12**, 737462.
