---
title: "Extracting and Padding Field Trial Layouts"
subtitle: "`padTrial()` and `plot_padTrial()`"
author: "biomAid package"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
```

---

## Overview

Field trials commonly contain a mix of plot types — test lines (e.g. DH or
breeding lines), check varieties, border plots, and guard rows — arranged
across one or more blocks. Before fitting a spatial model it is often
necessary to isolate the rectangular sub-grid occupied by the target plot
type and ensure that every grid position within that rectangle is represented
in the data frame, even if some plots are physically absent.

`padTrial()` automates this preparation in four steps, applied independently
within each group (block):

1. **Duplicate check** — stops with an informative error if any
   Row × Column position appears more than once within a group.
2. **Bounding box** — identifies the minimum and maximum row and column
   coordinates of the target plot type (`match`) to define the
   extraction rectangle.
3. **Subset** — retains only those rows that fall inside the bounding box,
   dropping guard rows and checks that lie outside it.
4. **Pad** — detects any Row × Column cells absent from the bounding box
   and inserts placeholder rows for them, writing `fill_value` into
   character/factor columns and `NA` into numeric columns.

An `add` column flags every row as `"old"` (original) or `"new"` (inserted).

`plot_padTrial()` visualises the result as a **Before / After** tile map,
with inserted cells shown in light grey.

---

## Argument reference

### `padTrial()`

| Argument | Default | Description |
|----------|---------|-------------|
| `data` | — | Input data frame containing the trial layout |
| `pattern` | `"Row:Column"` | Colon-separated names of the two spatial coordinate columns |
| `match` | `"DH"` | Plot type value(s) that define the target sub-trial |
| `split` | `"Block"` | Column(s) defining independent processing groups; `NULL` = whole dataset |
| `pad` | `TRUE` | Insert placeholder rows for missing grid cells |
| `keep` | `split` | Columns whose values are carried into padded rows |
| `fill_value` | `"Blank"` | String written into character/factor columns of padded rows |
| `type_col` | `"Type"` | Name of the plot-type column |
| `verbose` | `FALSE` | Print per-group bounding box and padding summary |

### `plot_padTrial()`

| Argument | Default | Description |
|----------|---------|-------------|
| `result` | — | Data frame returned by `padTrial()` |
| `data` | `NULL` | Original data; when supplied the Before panel includes guard rows |
| `type_col` | `"Type"` | Column used to colour tiles |
| `pattern` | `"Row:Column"` | Colon-separated spatial coordinate column names |
| `split` | `NULL` | Grouping column(s) — groups appear as facet columns |
| `label` | `NULL` | Column whose values are printed inside tiles |
| `theme` | `theme_bw()` | ggplot2 theme |
| `return_data` | `FALSE` | Return the tidy data frame instead of the plot |

---

## Example 1: Basic rectangular block with guard rows

The simplest case — a single rectangular block of test lines (DH) enclosed
by guard rows on all four sides. `padTrial()` detects the bounding box of
the DH plots, drops the guards, and (in this complete case) inserts no
padding.

```{r ex1_data}
# DH core: rows 3-8, columns 2-7  (6 rows × 6 cols = 36 DH plots)
# Guard border: row 1-2, row 9-10, col 1, col 8  (all surrounding cells)
set.seed(1L)
rows <- 1:10
cols <- 1:8

ex1 <- expand.grid(Row = rows, Column = cols, KEEP.OUT.ATTRS = FALSE)
ex1$Type  <- "Guard"
ex1$Block <- "B1"
ex1$Geno  <- NA_character_

# Assign DH plots to the core
core <- ex1$Row >= 3L & ex1$Row <= 8L & ex1$Column >= 2L & ex1$Column <= 7L
ex1$Type[core] <- "DH"
ex1$Geno[core] <- paste0("G", sprintf("%02d", seq_len(sum(core))))
```

```{r ex1_pad}
res1 <- padTrial(
  ex1,
  pattern  = "Row:Column",
  match    = "DH",
  split    = "Block",
  verbose  = TRUE
)

table(res1$add)
```

The Before panel (left, with `data =` supplied) shows the full layout
including guard rows. The After panel shows just the extracted DH core — no
padding was needed because the block was complete.

```{r ex1_plot}
#| fig-height: 5
p1 <- plot_padTrial(res1, data = ex1, split = "Block",
                    type_col = "Type", pattern = "Row:Column")
print(p1)
```

---

## Example 2: Block with internal missing plots

Real trials frequently have plots that could not be sown, were lost to
damage, or were deliberately left empty. Here several DH plots are absent
from the interior of the block. `padTrial()` detects the gaps and inserts
`"Blank"` placeholder rows so the grid is complete and rectangular — a
requirement for most spatial modelling software.

```{r ex2_data}
set.seed(2L)
ex2 <- expand.grid(Row = 1:8, Column = 1:6, KEEP.OUT.ATTRS = FALSE)
ex2$Type  <- "DH"
ex2$Block <- "B1"
ex2$Geno  <- paste0("G", sprintf("%03d", seq_len(nrow(ex2))))

# Remove 6 interior plots to simulate missing observations
missing_idx <- which(
  (ex2$Row == 2L & ex2$Column == 3L) |
  (ex2$Row == 4L & ex2$Column == 2L) |
  (ex2$Row == 4L & ex2$Column == 5L) |
  (ex2$Row == 5L & ex2$Column == 4L) |
  (ex2$Row == 6L & ex2$Column == 1L) |
  (ex2$Row == 7L & ex2$Column == 6L)
)
ex2 <- ex2[-missing_idx, ]

cat("Rows before padding:", nrow(ex2),
    "\nExpected after padding:", 8L * 6L, "\n")
```

```{r ex2_pad}
res2 <- padTrial(
  ex2,
  match      = "DH",
  split      = "Block",
  fill_value = "Blank",
  verbose    = TRUE
)

cat("'new' rows inserted:", sum(res2$add == "new"), "\n")
subset(res2, add == "new")[, c("Row", "Column", "Type", "Geno", "add")]
```

The Before panel shows the six white gaps where plots are missing. The After
panel shows the completed grid with inserted cells in light grey.

```{r ex2_plot}
#| fig-height: 5
p2 <- plot_padTrial(res2, data = ex2, split = "Block")
print(p2)
```

---

## Example 3: Non-rectangular block

Not all trial blocks are perfect rectangles. A common scenario is an
end-of-field block where some rows are shorter than others — the block is
trapezoidal or L-shaped. `padTrial()` computes the bounding box of all
target plots and pads the missing corner and edge positions to produce a
rectangular grid.

Here we construct a block shaped like a staircase: each successive pair of
rows has one fewer column on the right-hand side.

```{r ex3_data}
# Staircase layout:
#   Rows 1-2: columns 1-8
#   Rows 3-4: columns 1-6  (right 2 cols absent)
#   Rows 5-6: columns 1-4  (right 4 cols absent)
#   Rows 7-8: columns 1-2  (right 6 cols absent)

make_staircase <- function() {
  parts <- list(
    expand.grid(Row = 1:2, Column = 1:8),
    expand.grid(Row = 3:4, Column = 1:6),
    expand.grid(Row = 5:6, Column = 1:4),
    expand.grid(Row = 7:8, Column = 1:2)
  )
  d        <- do.call(rbind, parts)
  d$Type   <- "DH"
  d$Block  <- "B1"
  d$Geno   <- paste0("G", sprintf("%03d", seq_len(nrow(d))))
  d
}

ex3 <- make_staircase()
cat("Original rows:", nrow(ex3),
    "\nBounding box would be 8 rows × 8 cols =", 8L * 8L, "cells\n")
```

```{r ex3_pad}
res3 <- padTrial(
  ex3,
  match      = "DH",
  split      = "Block",
  fill_value = "Blank",
  verbose    = TRUE
)

cat("'new' rows inserted:", sum(res3$add == "new"), "\n")
```

The Before panel clearly shows the staircase shape — the missing corner
positions appear as white space. The After panel fills them all in with
`"Blank"` placeholder tiles (light grey) to produce a complete 8 × 8 grid.

```{r ex3_plot}
#| fig-height: 5
p3 <- plot_padTrial(res3, data = ex3, split = "Block")
print(p3)
```

---

## Example 4: Multiple blocks — each processed independently

When a trial has multiple blocks, `split` ensures that each block is
processed with its own bounding box. This is critical: blocks often occupy
different row/column ranges and have different numbers of missing plots.
Mixing them into a single pass would produce the wrong bounding box.

Here we build three blocks that differ in both shape and completeness:

- **Block B1** — complete 5 × 4 rectangle, no padding needed
- **Block B2** — 6 × 5 rectangle with 4 internal missing plots
- **Block B3** — non-rectangular (right-truncated), 5 rows × up to 6 cols

```{r ex4_data}
set.seed(4L)

# B1: complete 5×4
b1 <- expand.grid(Row = 1:5, Column = 1:4, KEEP.OUT.ATTRS = FALSE)
b1$Type  <- "DH"
b1$Block <- "B1"
b1$Geno  <- paste0("G", sprintf("%03d", seq_len(nrow(b1))))

# B2: 6×5 with 4 missing interior plots
b2 <- expand.grid(Row = 1:6, Column = 1:5, KEEP.OUT.ATTRS = FALSE)
b2$Type  <- "DH"
b2$Block <- "B2"
b2$Geno  <- paste0("G", sprintf("%03d", seq_len(nrow(b2))))
miss_b2  <- which(
  (b2$Row == 2L & b2$Column == 4L) |
  (b2$Row == 3L & b2$Column == 2L) |
  (b2$Row == 5L & b2$Column == 3L) |
  (b2$Row == 6L & b2$Column == 5L)
)
b2 <- b2[-miss_b2, ]

# B3: non-rectangular — right side truncated
b3_parts <- list(
  expand.grid(Row = 1:2, Column = 1:6),
  expand.grid(Row = 3:4, Column = 1:4),
  expand.grid(Row = 5:5, Column = 1:2)
)
b3        <- do.call(rbind, b3_parts)
b3$Type   <- "DH"
b3$Block  <- "B3"
b3$Geno   <- paste0("G", sprintf("%03d", seq_len(nrow(b3))))

ex4 <- rbind(b1, b2, b3)

cat("Rows per block (before padding):\n")
print(table(ex4$Block))
```

```{r ex4_pad}
res4 <- padTrial(
  ex4,
  match      = "DH",
  split      = "Block",
  fill_value = "Blank",
  verbose    = TRUE
)

cat("\nPadded rows per block:\n")
print(table(res4$Block[res4$add == "new"]))
```

With `split = "Block"` each block gets its own bounding box and its own
padding pass. The plot shows Before / After side-by-side for all three blocks.

```{r ex4_plot}
#| fig-height: 6
#| fig-width: 10
p4 <- plot_padTrial(res4, data = ex4, split = "Block")
print(p4)
```

---

## Example 5: Mixed plot types — DH lines and checks

Trials often contain both test lines and check varieties within the same
block. `padTrial()` uses `match` to define the bounding box from the target
type only, but retains **all** non-guard plots that fall inside that
bounding box — including checks.

```{r ex5_data}
set.seed(5L)
ex5 <- expand.grid(Row = 1:8, Column = 1:6, KEEP.OUT.ATTRS = FALSE)
ex5$Block <- "B1"

# Assign plot types: checks occupy every 6th position; guards on rows 1 & 8
ex5$Type  <- "DH"
check_idx <- seq(3L, nrow(ex5), by = 6L)
ex5$Type[check_idx]        <- "Check"
ex5$Type[ex5$Row == 1L]    <- "Guard"
ex5$Type[ex5$Row == 8L]    <- "Guard"

ex5$Geno <- ifelse(ex5$Type == "Guard", NA_character_,
                   paste0(ex5$Type, sprintf("%02d", seq_len(nrow(ex5)))))

# Remove 3 DH interior plots
miss5 <- which(ex5$Type == "DH")[c(2L, 10L, 18L)]
ex5   <- ex5[-miss5, ]
```

```{r ex5_pad}
res5 <- padTrial(
  ex5,
  match    = "DH",          # bounding box from DH only
  split    = "Block",
  verbose  = TRUE
)

cat("\nPlot types in result:\n")
print(table(res5$Type))
```

Guards on rows 1 and 8 are outside the DH bounding box and are dropped.
Checks inside the bounding box are retained. The three missing DH positions
are padded.

```{r ex5_plot}
#| fig-height: 5
p5 <- plot_padTrial(res5, data = ex5, split = "Block")
print(p5)
```

---

## Example 6: Multi-column split

When the trial spans multiple sites and blocks, `split` accepts a character
vector. Each unique combination of the split columns is processed
independently. The temporary composite key is dropped from the result —
only the original columns are returned.

```{r ex6_data}
# Two sites × two blocks each, some blocks with missing plots
make_site_block <- function(site, block, miss_rows = NULL, miss_cols = NULL) {
  d       <- expand.grid(Row = 1:5, Column = 1:4, KEEP.OUT.ATTRS = FALSE)
  d$Site  <- site
  d$Block <- block
  d$Type  <- "DH"
  d$Geno  <- paste0(site, block, "_G", sprintf("%02d", seq_len(nrow(d))))
  if (!is.null(miss_rows)) {
    rem <- which(d$Row %in% miss_rows & d$Column %in% miss_cols)
    if (length(rem)) d <- d[-rem, ]
  }
  d
}

ex6 <- rbind(
  make_site_block("S1", "B1"),
  make_site_block("S1", "B2", miss_rows = c(2L, 4L), miss_cols = c(3L)),
  make_site_block("S2", "B1", miss_rows = c(3L),     miss_cols = c(1L, 4L)),
  make_site_block("S2", "B2")
)
```

```{r ex6_pad}
res6 <- padTrial(
  ex6,
  match  = "DH",
  split  = c("Site", "Block"),
  keep   = c("Site", "Block"),
  verbose = TRUE
)

cat("\nPadded rows per Site × Block:\n")
with(res6[res6$add == "new", ], table(Site, Block))
```

---

## Extracting plot data — `return_data = TRUE`

`plot_padTrial()` with `return_data = TRUE` returns the tidy data frame used
internally to build the tile map. Columns are standardised: `row_num`,
`col_num`, `fill_group`, `panel` (`"Before"` / `"After"`), `group`, `is_new`,
and `label_text`.

```{r return_data}
df <- plot_padTrial(res2, data = ex2, split = "Block", return_data = TRUE)
head(df)
```

The data frame can be passed into any `ggplot2` workflow:

```{r bespoke}
#| fig-height: 4
df_aft <- df[df$panel == "After", ]

ggplot(df_aft, aes(x = col_num, y = row_num, fill = fill_group)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_tile(data = df_aft[df_aft$is_new, ],
            fill = "tomato", colour = "white", linewidth = 0.4) +
  scale_y_reverse() +
  scale_fill_manual(values = c(DH = "#4E79A7", Blank = "#DEDEDE"),
                    name = "Type") +
  labs(title = "Custom colour for padded cells",
       x = "Column", y = "Row") +
  theme_bw()
```

The ggplot object returned by `plot_padTrial()` can also be extended directly
with `+`:

```{r extend}
#| fig-height: 5
plot_padTrial(res4, data = ex4, split = "Block") +
  ggplot2::ggtitle("Three-block trial: before and after padding",
                   subtitle = "B1 complete · B2 internal gaps · B3 non-rectangular")
```

---

## Summary

| Task | Specification |
|------|---------------|
| Basic extraction (single block) | `padTrial(data, split = "Block")` |
| No blocking column | `split = NULL` |
| Multi-environment split | `split = c("Site", "Block"), keep = c("Site", "Block")` |
| Custom target plot type | `match = "Line"` (or any value in `type_col`) |
| Custom type column name | `type_col = "PlotKind"` |
| Custom fill string | `fill_value = "Missing"` |
| Suppress padding (extract only) | `pad = FALSE` |
| Per-group diagnostics | `verbose = TRUE` |
| Count padded rows | `table(result$add)` |
| Inspect padded rows | `subset(result, add == "new")` |
| Before/After plot (reconstruct Before) | `plot_padTrial(result)` |
| Before/After plot (full original layout) | `plot_padTrial(result, data = original)` |
| Multi-block faceted plot | `plot_padTrial(result, data = original, split = "Block")` |
| Label tiles with genotype | `label = "Geno"` |
| Export plot data | `return_data = TRUE` |
