---
title: "Random Regression of Treatment BLUPs from Multi-Treatment MET Analyses"
subtitle: "`randomRegress()` and `plot_randomRegress()`"
author: "Julian Taylor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 9
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
asreml_ok <- requireNamespace("asreml", quietly = TRUE)
if (asreml_ok) {
  library(asreml)
  library(ASExtras4)
  asreml.options(trace = FALSE)
}
```

---

## Overview

`randomRegress()` decomposes variety BLUPs from a multi-environment trial
into two biologically interpretable components:

- **Efficiency** — a variety's genetic value under the baseline (typically
  low-input) treatment.
- **Responsiveness (or Tolerance)** — a variety's capacity to respond
  genetically to the application of a treatment.

The decomposition is derived from the **G-matrix** (genetic
variance–covariance matrix across Treatment × Site combinations) using the
Gaussian conditional distribution. It generalises naturally from two
treatments to any number, with four built-in conditioning schemes.

`plot_randomRegress()` visualises the result as a regression scatter plot,
an efficiency–responsiveness quadrant plot, or a G-matrix heatmap.

---

## Mathematical Framework

### The multi-treatment, multi-environment LMM

Consider $n$ varieties evaluated across $s = 1, \ldots, S$ sites under
$t = 1, \ldots, T$ treatments. For variety $v$ at site $s$ under treatment $t$,
the linear mixed model is

$$
y_{tsv} = \tau_{ts} + u_{tsv} + \rho_{sr} + \varepsilon_{tsv},
$$

where $\tau_{ts}$ is a fixed Treatment × Site mean, $\rho_{sr}$ is a random
replicate-within-site effect, and $\varepsilon_{tsv}$ is the residual. The
genetic effect $u_{tsv}$ is the key quantity of interest.

Stacking observations across all treatment × site combinations into the
vector $\mathbf{y}$, the model can be written compactly as

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\tau} + \mathbf{Z}\mathbf{u} + \mathbf{W}\boldsymbol{\rho} + \boldsymbol{\varepsilon},
$$

where $\mathbf{X}$, $\mathbf{Z}$, and $\mathbf{W}$ are incidence matrices for
fixed effects, genetic effects, and replicates respectively.

**Genetic effects.** The $n \times (T \cdot S)$ matrix
$\mathbf{U} = [\mathbf{u}_{11} \mid \cdots \mid \mathbf{u}_{TS}]$ of
variety BLUPs stacked across all treatment × site groups satisfies

$$
\operatorname{vec}(\mathbf{U}) \sim \mathcal{N}(\mathbf{0},\; \mathbf{G} \otimes \mathbf{I}_n),
$$

where $\mathbf{G}$ is the $(T \cdot S) \times (T \cdot S)$ genetic
variance–covariance matrix (the G-matrix), and $\mathbf{I}_n$ is the
$n \times n$ identity over varieties. Each block $\mathbf{G}_{ts,t's'}$
describes the genetic covariance between treatment $t$ at site $s$ and
treatment $t'$ at site $s'$.

**Replicate variance.** The site-level replicate effects are modelled as
$\boldsymbol{\rho} \sim \mathcal{N}(\mathbf{0},\; \bigoplus_s \sigma^2_s \mathbf{I}_{r_s})$,
corresponding to the `at(Site):Rep` term in ASReml-R.

**Residual variance.** Within each site a separable first-order
autoregressive structure captures spatial trends along rows and columns,
$\boldsymbol{\varepsilon}_s \sim \mathcal{N}(\mathbf{0},\; \sigma^2_{e,s}\,\mathbf{A}_s(\rho_r) \otimes \mathbf{A}_s(\rho_c))$,
corresponding to `ar1(Row):ar1(Column)` nested within site.

For full derivations of the incidence matrices, Kronecker products, and
auto-generated genetic structures, see the
[Simulating Multi-Environment Trials](simTrialData.html) vignette.

In ASReml-R V4 this model is written as:

```r
model <- asreml(
  yield    ~ TSite,
  random   = ~ corgh(TSite):Variety + at(Site):Rep,
  residual = ~ dsum(~ ar1(Row):ar1(Column) | Site),
  data     = dat
)
```

where `TSite` is a composite factor formed by pasting the treatment label
onto the site label (e.g. `"T0-Site1"`). This is necessary because
ASReml-R requires a single factor to index the columns of the G-matrix.

---

### The G-matrix

The G-matrix returned by ASReml-R V4 for the `struct(TSite):Variety` random
term is indexed by the levels of `TSite`. For a single site with three
treatments, $\mathbf{G}$ is a $3 \times 3$ matrix; for four sites and three
treatments it is $12 \times 12$.

The structure keyword (`corgh`, `us`, `fa`, `diag`, …) controls how
$\mathbf{G}$ is parameterised. `randomRegress()` accepts any of these:

| ASReml-R term | G-matrix form |
|---|---|
| `corgh(TSite):Variety` | Heterogeneous variances, one shared correlation parameter per pair |
| `corh(TSite):Variety` | Heterogeneous variances, single correlation (2-group) |
| `us(TSite):Variety` | Fully unstructured covariance matrix |
| `fa(TSite, k):Variety` | Factor analytic of order $k$ |
| `diag(TSite):Variety` | Diagonal — zero between-group covariances |

---

### Conditional decomposition

For treatment-site combination $j$ with conditioning set $A_j$ (a subset of
the other treatment-site labels), the **multivariate conditional normal**
distribution gives the regression of $\hat{u}_{j}$ on $\hat{\mathbf{u}}_{A_j}$:

$$
\hat{u}_{j} \mid \hat{\mathbf{u}}_{A_j}
  \sim \mathcal{N}\!\left(
    \boldsymbol{\beta}_j^\top \hat{\mathbf{u}}_{A_j},\;
    \sigma^2_{j \mid A_j}
  \right),
$$

where the **regression coefficients** and **conditional genetic variance** are

$$
\boldsymbol{\beta}_j = \mathbf{G}_{A_j A_j}^{-1}\, \mathbf{G}_{A_j j},
\qquad
\sigma^2_{j \mid A_j} = G_{jj} - \mathbf{G}_{j A_j}\,\boldsymbol{\beta}_j.
$$

The **responsiveness BLUP** for variety $v$ is the residual from this
regression:

$$
\tilde{u}_{j,v} = \hat{u}_{j,v} - \boldsymbol{\beta}_j^\top \hat{\mathbf{u}}_{A_j,v}.
$$

Responsiveness BLUPs are:

- **zero-mean** within each treatment-site group (inherited from the BLUPs)
- **orthogonal** to the conditioning BLUPs in the G-matrix sense
- **genetically independent** of the efficiency signal, with variance
  $\sigma^2_{j \mid A_j}$

---

### The transformation matrix

Collecting all conditioning relationships defines a sparse transformation
matrix $\mathbf{T}$ such that:

$$
\tilde{\mathbf{u}} = \mathbf{T}\hat{\mathbf{u}},
\qquad
\operatorname{Var}(\tilde{\mathbf{u}}) = \mathbf{T}\mathbf{G}\mathbf{T}^\top
  \equiv \mathbf{V}_T.
$$

The diagonal blocks of $\mathbf{V}_T$ are the conditional genetic variances
$\sigma^2_{j \mid A_j}$. For `type = "sequential"` the full
$\mathbf{V}_T$ is **diagonal** — a complete Gram-Schmidt
orthogonalisation of the BLUP space.

---

### Conditioning schemes

The four built-in schemes differ only in how $A_j$ is chosen:

| Scheme | $A_j$ for treatment $j$ | $\mathbf{V}_T$ structure | Best for |
|--------|------------------------|--------------------------|----------|
| `"baseline"` | $\{\text{levs}[1]\}$ only | Block-diagonal | Comparing all treatments to one baseline |
| `"sequential"` | $\{\text{levs}[1], \ldots, \text{levs}[j-1]\}$ | Diagonal ($LDL^\top$ Cholesky) | Fully orthogonal components |
| `"partial"` | All other treatments simultaneously | Dense | Partial genetic variances |
| `"custom"` | User-specified | Depends | Bespoke experimental logic |

---

## Example 1 — Single site, partial conditioning

> **Requires:** ASReml-R V4 and ASExtras4. The vignette is pre-rendered so
> no licence is needed to read it; run the code chunks only if you have
> both packages installed.

### Simulated data

We simulate a single-site trial with 50 varieties and three nitrogen
treatments (T0 = control, T1 = low N, T2 = high N), three replicates per
treatment. With only one site, the G-matrix is a $3 \times 3$ matrix
describing the genetic (co)variances between treatments — a clean,
interpretable structure for introducing the method.

We use `sim.options` to restrict the auto-generated between-treatment
genetic correlations to a moderate range (0.55–0.70). This keeps the true
genetic structure well away from the parameter boundary, giving the
`corgh` model a stable likelihood surface to optimise. The default
`sigma_genetic` and `error_sd` values produce a per-plot heritability
of approximately 0.60 — ample signal for reliable estimation with 50
varieties and 3 replicates.

```{r sim1}
#| cache: true
out1 <- simTrialData(
  nvar       = 50L,
  nsite      = 1L,
  treatments = c("T0", "T1", "T2"),
  nrep       = 3L,
  seed       = 7L,
  verbose    = FALSE,
  sim.options = list(g_cor_min = 0.55, g_cor_max = 0.70)
)
dat1 <- out1$data
```

The field layout and true genetic correlation structure can be examined
with `plot_simTrialData()`:

```{r plot_sim1_layout}
#| fig-height: 4
plot_simTrialData(out1, type = "trial", fill = "Treatment", ncol = 3L)
```

```{r plot_sim1_corr}
#| fig-height: 4
#| fig-width: 5
plot_simTrialData(out1, type = "correlation")
```

The correlation heatmap confirms that the three treatments have positive
genetic correlations with each other (true values in the range 0.55–0.70),
reflecting that varieties with good baseline performance also tend to respond
positively to applied nitrogen — but not perfectly. This partial sharing of
genetic rankings is exactly the structure that partial conditioning is
designed to dissect.

### Forming `TSite` and fitting the model

ASReml-R requires a single factor to index the G-matrix columns. We create
`TSite` by pasting the treatment label onto the site name. With only one
site this simply appends the site label (e.g. `"T0-Site1"`), but the same
code generalises directly to multi-site data:

```{r prep1}
dat1$TSite  <- interaction(dat1$Treatment, dat1$Site, sep = "-")
dat1$Row    <- factor(dat1$Row)
dat1$Column <- factor(dat1$Column)
levs1       <- c("T0", "T1", "T2")
```

We fit a **heterogeneous-correlation** (`corgh`) model. With three
treatments this parameterises the $3 \times 3$ G-matrix with three
genetic variances and three pairwise correlations — the most flexible
fully-identified structure for this problem size:

```{r fit1}
#| eval: !expr asreml_ok
#| cache: true
#| results: hide
model1 <- asreml(
  yield    ~ TSite,
  random   = ~ corgh(TSite):Variety + at(Site):Rep,
  residual = ~ ar1(Row):ar1(Column),
  data     = dat1,
  maxit    = 30L
)
model1 <- update(model1)
```

```{r vc1}
#| eval: !expr asreml_ok
#| dependson: "fit1"
vc1 <- summary(model1)$varcomp
vc1[grepl("TSite", rownames(vc1)), ]
```

The `corgh` varcomp rows give the genetic variance for each `TSite` level
(diagonal of the G-matrix) and the estimated genetic correlation between
each pair. With true correlations in the range 0.55–0.70 and 50 varieties
across 3 replicates, the estimates should be well away from the boundary
($|r| \ll 1$), ensuring the reconstructed G-matrix is positive definite and
all conditional variances are positive.

### Running `randomRegress()` — partial conditioning

With a single site, **partial conditioning** is the natural choice: each
treatment is conditioned on all the others simultaneously, so the three
responsiveness BLUPs capture the genetic signal that is unique to each
treatment and unshared with the others.

```{r rr1}
#| eval: !expr asreml_ok
#| cache: true
#| dependson: "fit1"
res1 <- randomRegress(
  model1,
  term = "corgh(TSite):Variety",
  levs = levs1,
  type = "partial",
  sep  = "-"
)
```

With a single site, `beta` has one row and `sigmat` has one row — the
regression coefficients and conditional genetic variance are scalars:

```{r show_beta1}
#| eval: !expr asreml_ok
#| dependson: "rr1"
# Site-specific regression coefficients (one site -> one row each)
res1$beta
```

```{r show_sigmat1}
#| eval: !expr asreml_ok
#| dependson: "rr1"
# Conditional genetic variances under partial conditioning
res1$sigmat
```

Each `sigmat` entry $\sigma^2_{j \mid A_j}$ is the genetic variance
in treatment $j$ that is **orthogonal** to all other treatments. A small
value relative to $G_{jj}$ (the diagonal of `Gmat`) indicates that most of
the genetic variation in that treatment is shared with the others;
a value close to $G_{jj}$ indicates a large treatment-specific genetic
component. **Negative values** would indicate a near-singular (indefinite)
G-matrix caused by boundary correlation estimates — `randomRegress()` warns
explicitly when this occurs and sets the affected entries to `NA`.

```{r compare_gvar1}
#| eval: !expr asreml_ok
#| dependson: "rr1"
# Compare conditional variances to marginal variances
data.frame(
  treatment  = levs1,
  marginal   = round(diag(res1$Gmat), 4),
  conditional = round(res1$sigmat[1L, ], 4)
)
```

### Regression plot

The regression plot shows the raw treatment BLUP (y-axis) against the
conditioning BLUP (x-axis). Under partial conditioning the conditioning
set includes **all other treatments**, so there is a single regression for
each pair. The slope $\hat{\beta}$ is annotated in the top-left corner.

```{r plot_reg1}
#| eval: !expr asreml_ok
#| fig-height: 5
#| dependson: "rr1"
p_reg1 <- plot_randomRegress(res1, type = "regress")
print(p_reg1)
```

A slope $\hat{\beta} > 0$ and close to 1 means the two treatments share
similar genetic rankings; a slope close to 0 means the treatment adds
little information once the others are accounted for.

### Quadrant plot with default highlights

The quadrant plot places **efficiency** on the x-axis and
**responsiveness** on the y-axis. Under partial conditioning, every
treatment has a responsiveness axis, so there are three quadrant panels:

```{r plot_quad1}
#| eval: !expr asreml_ok
#| fig-height: 7
#| dependson: "rr1"
p_quad1 <- plot_randomRegress(res1, type = "quadrant", highlight = "default")
print(p_quad1)
```

The default highlight algorithm selects three archetypes from the top-right
quadrant (high efficiency + high responsiveness) and three from the
bottom-left (low efficiency + low responsiveness), annotated consistently
across all panels. Varieties that appear in the top-right of all panels
respond positively to every treatment; those that shift between quadrants
across panels show treatment-specific responses.

### G-matrix heatmap

With a single site the G-matrix is a $3 \times 3$ correlation matrix —
straightforward to interpret:

```{r plot_gmat1}
#| eval: !expr asreml_ok
#| fig-height: 4
#| fig-width: 5
#| dependson: "rr1"
p_gmat1 <- plot_randomRegress(res1, type = "gmat")
print(p_gmat1)
```

The heatmap displays `cov2cor(Gmat)`. High off-diagonal values (dark red)
indicate strong genetic co-ranking between treatments; values below ~0.8
suggest meaningful genotype × treatment interaction that the partial
decomposition has captured.

---

## Example 2 — Multi-site, factor analytic model

### Simulated data

We now scale to a multi-environment setting: 20 varieties, 4 sites, and the
same three nitrogen treatments. The FA(2) model is the standard choice for
MET analyses — it captures the dominant axes of genotype × environment
interaction parsimoniously.

```{r sim2}
#| cache: true
out2 <- simTrialData(
  nvar       = 20L,
  nsite      = 4L,
  treatments = c("T0", "T1", "T2"),
  nrep       = 3L,
  seed       = 42L,
  verbose    = FALSE
)
dat2 <- out2$data
```

The `"blup"` plot of `plot_simTrialData()` visualises the true genetic
values across all 12 treatment × site groups, revealing the GEI structure
baked into the simulation:

```{r plot_sim2_blup}
#| fig-height: 5
#| dependson: "sim2"
plot_simTrialData(out2, type = "blup", sort = TRUE)
```

Varieties cluster into those with broadly positive responses across all
treatment × site groups (top rows) and those with more variable or
treatment-specific patterns — exactly the structure that `randomRegress()`
is designed to unpack.

### Forming `TSite` and fitting the FA(2) model

```{r prep2}
#| dependson: "sim2"
dat2$TSite  <- interaction(dat2$Treatment, dat2$Site, sep = "-")
dat2$Row    <- factor(dat2$Row)
dat2$Column <- factor(dat2$Column)
levs2       <- c("T0", "T1", "T2")
```

A **factor analytic** model with 2 factors provides a parsimonious
description of the $12 \times 12$ G-matrix across 4 sites and 3 treatments.
The two factors capture the dominant axes of genetic co-variation —
typically a general adaptation factor and a GEI factor:

```{r fit2}
#| eval: !expr asreml_ok
#| cache: true
#| results: hide
model2 <- asreml(
  yield    ~ TSite,
  random   = ~ fa(TSite, 2):Variety + at(Site):Rep,
  residual = ~ dsum(~ ar1(Row):ar1(Column) | Site),
  data     = dat2,
  maxit    = 30L
)
model2 <- update(model2)
```

```{r vc2}
#| eval: !expr asreml_ok
#| dependson: "fit2"
vc2 <- summary(model2)$varcomp
vc2[grepl("TSite", rownames(vc2)), ]
```

The varcomp rows show the factor loadings (`!fa1`, `!fa2`) and specific
variances (`!var`) for each `TSite` level. The G-matrix reconstructed from
these is $\hat{\mathbf{G}} = \hat{\boldsymbol{\Lambda}}\hat{\boldsymbol{\Lambda}}^\top + \hat{\boldsymbol{\Psi}}$.

### Running `randomRegress()` — baseline conditioning

For multi-site data with an FA model, **baseline conditioning** is the most
widely used scheme: T0 is the efficiency (control) treatment, and T1 and T2
are conditioned on T0 to give treatment-specific responsiveness BLUPs.

```{r rr2}
#| eval: !expr asreml_ok
#| cache: true
#| dependson: "fit2"
res2 <- randomRegress(
  model2,
  term = "fa(TSite, 2):Variety",
  levs = levs2,
  type = "baseline",
  sep  = "-"
)
```

With 4 sites, `beta` has 4 rows (one per site) and `sigmat` has 4 rows.
The site-specific regression slopes in `beta` capture how the treatment
response relationship changes across environments:

```{r show_beta2}
#| eval: !expr asreml_ok
#| dependson: "rr2"
# Site-specific regression coefficients for T1 conditioned on T0
round(res2$beta$T1, 3)
```

### Regression plot

```{r plot_reg2}
#| eval: !expr asreml_ok
#| fig-height: 7
#| dependson: "rr2"
p_reg2 <- plot_randomRegress(res2, type = "regress")
print(p_reg2)
```

Each panel corresponds to one site × treatment pair. The dotted regression
line through the origin carries the site-specific $\hat{\beta}$ slope.
Panels where $\hat{\beta}$ differs substantially across sites indicate
heterogeneous treatment responses — the G-matrix captures this as
off-diagonal structure between treatment × site combinations.

### Quadrant plot — default highlights

```{r plot_quad2}
#| eval: !expr asreml_ok
#| fig-height: 7
#| dependson: "rr2"
p_quad2 <- plot_randomRegress(res2, type = "quadrant", highlight = "default")
print(p_quad2)
```

The automatic highlight algorithm identifies the six most informative
archetypes by their position in the efficiency–responsiveness space,
averaged across all sites. Orange points are top-right archetypes (broadly
adapted and responsive); blue points are bottom-left archetypes (consistently
poor performers). The same six varieties are labelled in every panel.

### Filtering to a single treatment pair

For multi-site multi-treatment results, `treatments` restricts which panels
are shown. This is useful when a particular treatment comparison is of
primary interest:

```{r plot_quad2_t1}
#| eval: !expr asreml_ok
#| fig-height: 5
#| dependson: "rr2"
p_t1 <- plot_randomRegress(res2, type = "quadrant",
                            treatments = "T1", highlight = "default")
print(p_t1)
```

### G-matrix heatmap

The $12 \times 12$ G-matrix correlation heatmap reveals the full structure
of genetic co-variation across all treatment × site combinations:

```{r plot_gmat2}
#| eval: !expr asreml_ok
#| fig-height: 6
#| fig-width: 8
#| dependson: "rr2"
p_gmat2 <- plot_randomRegress(res2, type = "gmat")
print(p_gmat2)
```

Within-treatment correlations (e.g. T0-Env1 with T0-Env2) are typically
high, reflecting consistent genetic ranking for that treatment across sites.
Correlations across treatments (e.g. T0-Env1 with T2-Env3) reflect the
combined effect of genotype × treatment and genotype × environment interaction.

### Bespoke visualisation with `return_data = TRUE`

All three plot types support `return_data = TRUE`, which returns the tidy
data frame used to build the plot. This enables custom `ggplot2`
visualisations beyond the built-in types:

```{r return_data}
#| eval: !expr asreml_ok
#| fig-height: 4
#| dependson: "rr2"
df_t1 <- plot_randomRegress(res2, type = "quadrant",
                             treatments  = "T1",
                             return_data = TRUE)

ggplot(df_t1, aes(x = x, y = y, colour = Site)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_point(size = 1.8, alpha = 0.8) +
  facet_wrap(~ Site, nrow = 2L) +
  labs(x = "Efficiency (T0 BLUP)",
       y = "Responsiveness (T1 | T0)",
       title = "T1 responsiveness by site") +
  theme_bw() +
  theme(legend.position = "none")
```

---

## Return value summary

`randomRegress()` returns a named list:

| Element | Type | Description |
|---------|------|-------------|
| `blups` | Data frame | Raw BLUPs + responsiveness BLUPs + HSD per variety × site |
| `beta` | Named list | Site-specific $\boldsymbol{\beta}_j$ matrices, one per conditioned treatment |
| `Gmat` | Matrix | Original G-matrix (covariance) |
| `TGmat` | Matrix | Transformed G-matrix $\mathbf{T}\mathbf{G}\mathbf{T}^\top$ |
| `sigmat` | Matrix | Conditional genetic variances $\sigma^2_{j \mid A_j}$ per site |
| `tmat` | Matrix | Full transformation matrix $\mathbf{T}$ |
| `cond_list` | Named list | Resolved conditioning structure |
| `type` | Character | Conditioning scheme used |

---

## Summary

| Task | Code |
|------|------|
| Baseline conditioning | `type = "baseline"` (default) |
| Fully orthogonal (Gram-Schmidt) | `type = "sequential"` |
| Partial genetic variances | `type = "partial"` |
| Bespoke conditioning | `type = "custom", cond = list(...)` |
| Regression scatter plot | `plot_randomRegress(res, type = "regress")` |
| Efficiency–responsiveness plot | `plot_randomRegress(res, type = "quadrant")` |
| G-matrix heatmap | `plot_randomRegress(res, type = "gmat")` |
| Default archetype highlights | `highlight = "default"` |
| Named variety highlights | `highlight = c("Var01", "Var15")` |
| Suppress highlights | `highlight = NULL` |
| Restrict treatment panels | `treatments = "T1"` |
| Export plot data | `return_data = TRUE` |
| Absolute yield scale | `centre = TRUE` |
| Single site, corgh term | `term = "corgh(TSite):Variety"` |
| Multi-site, FA term | `term = "fa(TSite, 2):Variety"` |

---

## References

Lemerle, D., Smith, A., Verbeek, B., Koetz, E., Lockley, P. & Martin, P. (2006).
Incremental crop tolerance to weeds: A measure for selecting competitive ability
in Australian wheats. *Euphytica*, **149**, 85–95.
<https://doi.org/10.1007/s10681-005-9056-5>

McDonald G, Bovill W, Taylor J, Wheeler R. (2015) Responses to phosphorus among wheat genotypes.
*Crop and Pasture Science*, **66**(5), 430–444. <https://doi.org/10.1071/CP14191>
