---
title: "Multivariate Random Regression of Treatment BLUPs"
subtitle: "`randomRegress()` and `plot_randomRegress()`"
author: "biomAid package"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 9
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
```

---

## Overview

`randomRegress()` decomposes variety BLUPs from a multi-environment trial
into two biologically interpretable components:

- **Efficiency** — a variety's genetic value under the baseline (typically
  low-input) treatment, reflecting its ability to perform well without
  additional inputs.
- **Responsiveness** — a variety's capacity to respond **beyond** what its
  baseline genetic value predicts, reflecting the additional gain from
  intensification.

The decomposition is derived from the **G-matrix** (genetic
variance–covariance matrix across Treatment × Site combinations) using the
Gaussian conditional distribution. It generalises naturally from two
treatments to any number, with four built-in conditioning schemes.

`plot_randomRegress()` visualises the result as a regression scatter plot,
an efficiency–responsiveness quadrant plot, or a G-matrix heatmap.

---

## Mathematical Framework

### The G-matrix

Let $\hat{\mathbf{u}}_s = (\hat{u}_{1s}, \ldots, \hat{u}_{ks})^\top$ be the
vector of variety BLUPs for $k$ treatments at site $s$, stacked across all
sites. The genetic variance–covariance structure is described by the
**G-matrix**:

$$
\mathbf{G} = \operatorname{Var}\!\left(\hat{\mathbf{u}}\right),
$$

with $k \times k$ blocks $\mathbf{G}_{jl}$ giving the genetic covariance
between treatments $j$ and $l$ averaged across sites. In ASReml-R V4 this
is the variance parameter matrix for the Treatment-Site:Variety random term.

---

### Conditional decomposition

For treatment $j$ with conditioning set $A_j$ (a subset of the other
treatment labels), the **multivariate conditional normal** distribution gives
the regression of $\hat{u}_{js}$ on $\hat{\mathbf{u}}_{A_j,s}$:

$$
\hat{u}_{js} \mid \hat{\mathbf{u}}_{A_j,s}
  \sim \mathcal{N}\!\left(
    \boldsymbol{\beta}_j^\top \hat{\mathbf{u}}_{A_j,s},\;
    \sigma^2_{j \mid A_j}
  \right),
$$

where the **regression coefficients** and **conditional genetic variance** are

$$
\boldsymbol{\beta}_j = \mathbf{G}_{A_j A_j}^{-1}\, \mathbf{G}_{A_j j},
\qquad
\sigma^2_{j \mid A_j} = G_{jj} - \mathbf{G}_{j A_j}\,\boldsymbol{\beta}_j.
$$

The **responsiveness BLUP** for variety $v$ at site $s$ is the residual from
this regression:

$$
\tilde{u}_{js,v} = \hat{u}_{js,v} - \boldsymbol{\beta}_j^\top \hat{\mathbf{u}}_{A_j,s,v}.
$$

Responsiveness BLUPs are:

- **zero-mean** within each site (inherited from the BLUPs)
- **orthogonal** to the conditioning BLUPs in the G-matrix sense
- **genetically independent** of the efficiency signal, with variance
  $\sigma^2_{j \mid A_j}$

---

### The transformation matrix

Collecting all conditioning relationships across treatments and sites defines
a sparse lower-triangular transformation matrix $\mathbf{T}$ such that:

$$
\tilde{\mathbf{u}} = \mathbf{T}\hat{\mathbf{u}},
\qquad
\operatorname{Var}(\tilde{\mathbf{u}}) = \mathbf{T}\mathbf{G}\mathbf{T}^\top
  \equiv \mathbf{V}_T.
$$

The diagonal blocks of $\mathbf{V}_T$ are the conditional genetic variances
$\sigma^2_{j \mid A_j}$; off-diagonal blocks measure residual genetic
associations between components. For `type = "sequential"` the full
$\mathbf{V}_T$ is **diagonal** — a complete Gram-Schmidt
orthogonalisation of the BLUP space.

---

### Conditioning schemes

The four built-in schemes differ only in how $A_j$ is chosen:

| Scheme | $A_j$ for treatment $j$ | $\mathbf{V}_T$ structure | Best for |
|--------|------------------------|--------------------------|----------|
| `"baseline"` | $\{\text{levs}[1]\}$ only | Block-diagonal | Comparing all treatments to one baseline |
| `"sequential"` | $\{\text{levs}[1], \ldots, \text{levs}[j-1]\}$ | Diagonal ($LDL^\top$ Cholesky) | Fully orthogonal components |
| `"partial"` | All other treatments | Dense | Partial genetic variances |
| `"custom"` | User-specified | Depends | Bespoke experimental logic |

---

## Setup

### Simulated data

We use `simTrialData()` to generate a realistic multi-environment trial with
30 varieties, 6 sites, and 3 nitrogen treatments (T0 = control, T1 = low N,
T2 = high N). From the raw data we construct a mock `randomRegress()` result
object directly from within-site variety means — the same approach used in
practice when ASReml BLUPs are replaced by de-regressed or adjusted means.

```{r simulate}
set.seed(99L)
out <- simTrialData(
  nvar       = 30L,
  nsite      = 6L,
  treatments = c("T0", "T1", "T2"),
  nrep       = 3L,
  seed       = 99L,
  verbose    = FALSE
)
dat   <- out$data
sites <- levels(dat$Site)
levs  <- c("T0", "T1", "T2")
ns    <- length(sites)
nvar  <- 30L
```

```{r build_res}
#| code-fold: true
#| code-summary: "Build mock randomRegress() result (click to expand)"

# Within-site, within-treatment variety means (stand-ins for BLUPs)
blup_list <- lapply(sites, function(s) {
  sub <- dat[dat$Site == s, ]
  u0 <- tapply(sub$yield[sub$Treatment == "T0"],
               sub$Variety[sub$Treatment == "T0"], mean)
  u1 <- tapply(sub$yield[sub$Treatment == "T1"],
               sub$Variety[sub$Treatment == "T1"], mean)
  u2 <- tapply(sub$yield[sub$Treatment == "T2"],
               sub$Variety[sub$Treatment == "T2"], mean)
  # Centre within site (BLUPs are zero-mean)
  u0c <- u0 - mean(u0)
  u1c <- u1 - mean(u1)
  u2c <- u2 - mean(u2)
  # Baseline regression coefficients
  b1 <- cov(u0c, u1c) / var(u0c)
  b2 <- cov(u0c, u2c) / var(u0c)
  data.frame(
    Site    = s,
    Variety = names(u0c),
    T0      = as.numeric(u0c),
    T1      = as.numeric(u1c),
    T2      = as.numeric(u2c),
    resp.T1 = as.numeric(u1c - b1 * u0c),
    resp.T2 = as.numeric(u2c - b2 * u0c),
    HSD.T1  = NA_real_,
    HSD.T2  = NA_real_,
    stringsAsFactors = FALSE
  )
})
blups <- do.call(rbind, blup_list)

# Site-specific beta matrices
beta_T1 <- matrix(NA_real_, ns, 1L, dimnames = list(sites, "T0"))
beta_T2 <- matrix(NA_real_, ns, 1L, dimnames = list(sites, "T0"))
for (i in seq_along(sites)) {
  s <- sites[i]
  u0 <- blups$T0[blups$Site == s]
  u1 <- blups$T1[blups$Site == s]
  u2 <- blups$T2[blups$Site == s]
  beta_T1[i, "T0"] <- cov(u0, u1) / var(u0)
  beta_T2[i, "T0"] <- cov(u0, u2) / var(u0)
}

# G-matrix from variety means across all treatment x site combinations
nms <- paste0(rep(levs, each = ns), "-", rep(sites, length(levs)))
mat <- matrix(NA_real_, nrow(blups), length(nms),
              dimnames = list(NULL, nms))
for (nm in nms) {
  parts     <- strsplit(nm, "-")[[1L]]
  mat[, nm] <- blups[[parts[1L]]][blups$Site == parts[2L]]
}
Gmat <- cov(mat, use = "complete.obs")

# Assemble mock res object
res <- list(
  blups     = blups,
  beta      = list(T1 = beta_T1, T2 = beta_T2),
  Gmat      = Gmat,
  TGmat     = Gmat,
  sigmat    = matrix(runif(ns * 2L, 100, 400), ns, 2L,
                     dimnames = list(sites, c("T1", "T2"))),
  tmat      = diag(length(nms)),
  cond_list = list(T0 = NULL, T1 = "T0", T2 = "T0"),
  type      = "baseline"
)
```

The result object has the structure returned by `randomRegress()`:

```{r show_structure}
str(res, max.level = 1)
```

---

## Regression plot — `type = "regress"`

The regression plot shows the raw treatment BLUPs for each conditioned
treatment (y-axis) against the conditioning treatment BLUPs (x-axis). Each
panel corresponds to one site × treatment pair. A **dotted random regression
line** through the origin carries the site-specific slope $\hat{\beta}$,
annotated in the top-left corner.

The slope $\hat{\beta}$ quantifies the average relationship between
treatments: $\hat{\beta} > 1$ means the conditioned treatment varies more
than the baseline; $\hat{\beta} < 1$ means it varies less.

```{r plot_regress}
#| fig-height: 7
p_reg <- plot_randomRegress(res, type = "regress", centre = TRUE)
print(p_reg)
```

The `centre = TRUE` option adds back the within-site mean, placing BLUPs on
an approximate absolute yield scale. For true ASReml BLUPs (which are
already zero-mean by construction) `centre = FALSE` (the default) is
appropriate.

---

## Quadrant plot — `type = "quadrant"`

The quadrant plot separates the two components of genetic performance.
The **x-axis** carries the unconditional (efficiency) BLUPs; the **y-axis**
carries the responsiveness BLUPs $\tilde{u}_{js,v}$. Dotted zero lines on
both axes divide each panel into four quadrants:

| Quadrant | Efficiency | Responsiveness | Interpretation |
|----------|-----------|----------------|----------------|
| Top-right | High | High | Broadly adapted: performs well and responds strongly |
| Top-left | Low | High | Responsive but low-yielding under baseline |
| Bottom-right | High | Low | Efficient but doesn't benefit from intensification |
| Bottom-left | Low | Low | Poor performance on both scales |

```{r plot_quadrant}
#| fig-height: 7
p_quad <- plot_randomRegress(res, type = "quadrant")
print(p_quad)
```

---

## Variety highlighting

By default, `highlight = "default"` automatically selects and annotates
**six varieties** that are informative archetypes of the efficiency–
responsiveness space. The algorithm operates in the centred quadrant space,
independently of the `centre` argument used for the plot axes.

### How the default algorithm works

1. Compute the mean efficiency (x) and responsiveness (y) for each variety
   **across all sites and treatment pairs**.
2. Separate varieties into the **top-right** quadrant ($x > 0$, $y > 0$)
   and the **bottom-left** quadrant ($x < 0$, $y < 0$).
3. Within each quadrant, retain varieties whose distance from the origin
   exceeds the 50th percentile (relaxes to 33rd then 0th if fewer than
   three candidates remain).
4. Pick three archetypes by polar angle $\theta = \arctan(|y|/|x|)$:
   - **Efficiency archetype** — smallest $\theta$ (close to the x-axis)
   - **Balanced archetype** — closest to $\theta = \pi/4$
   - **Responsiveness archetype** — largest $\theta$ (close to the y-axis)

The same six varieties are annotated consistently across all site and
treatment-pair panels.

```{r highlight_default}
#| fig-height: 7
# Orange = top-right archetypes, Blue = bottom-left archetypes
p_hl <- plot_randomRegress(res, type = "quadrant", highlight = "default")
print(p_hl)
```

### User-specified highlights

Pass a character vector of variety names to `highlight` to annotate specific
varieties of interest (shown in green):

```{r highlight_custom}
#| fig-height: 7
user_vars <- levels(factor(res$blups$Variety))[c(1L, 10L, 20L)]
p_custom  <- plot_randomRegress(res, type  = "quadrant",
                                highlight = user_vars)
print(p_custom)
```

### No highlighting

Set `highlight = NULL` to suppress annotation entirely and draw all points
in a single colour:

```{r highlight_none}
#| fig-height: 7
p_none <- plot_randomRegress(res, type = "quadrant", highlight = NULL)
print(p_none)
```

---

## Filtering treatments — `treatments` argument

For results with many conditioned treatments, `treatments` restricts which
pairs appear in the plot:

```{r treatments_filter}
#| fig-height: 4
p_t1 <- plot_randomRegress(res, type = "quadrant", treatments = "T1")
print(p_t1)
```

---

## G-matrix heatmap — `type = "gmat"`

The `"gmat"` plot displays `cov2cor(Gmat)` — the genetic correlations
between all Treatment × Site combinations — as a diverging heatmap. High
positive correlations (dark red) indicate treatments and/or sites that share
genetic rankings; negative correlations (dark blue) indicate rank reversal.

```{r plot_gmat}
#| fig-height: 6
#| fig-width: 8
p_gmat <- plot_randomRegress(res, type = "gmat")
print(p_gmat)
```

Correlations within the same treatment across sites (e.g. T0-Env1 with
T0-Env2) tend to be moderate to high. Correlations between treatments
(e.g. T0-Env1 with T2-Env3) reflect the degree of Treatment × Genotype ×
Environment interaction.

---

## Extracting plot data — `return_data = TRUE`

All three plot types support `return_data = TRUE`, which returns the tidy
data frame used to build the plot. This is useful for bespoke
visualisations or downstream analysis.

```{r return_data}
df_quad <- plot_randomRegress(res, type = "quadrant", return_data = TRUE)
head(df_quad)
```

The data frame can be passed to any `ggplot2` workflow:

```{r bespoke}
#| fig-height: 4
# Custom plot: quadrant for T1 only, custom colour by site
df_t1 <- plot_randomRegress(res, type = "quadrant",
                             treatments  = "T1",
                             return_data = TRUE)

ggplot(df_t1, aes(x = x, y = y, colour = Site)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_point(size = 1.8, alpha = 0.8) +
  facet_wrap(~ Site, nrow = 2L) +
  labs(x = "Efficiency (T0 BLUP)", y = "Responsiveness (T1 | T0)",
       title = "Custom site-coloured quadrant plot") +
  theme_bw() +
  theme(legend.position = "none")
```

---

## ASReml-R V4 integration

In practice `randomRegress()` takes an ASReml-R V4 model object. The
workflow below requires a licence and is shown for reference only.

```{r asreml_workflow}
#| eval: false
library(asreml)

# Fit a factor analytic or unstructured covariance model
model <- asreml(
  yield  ~ Treatment + Site,
  random = ~ us(Treatment):Site:Variety,
  data   = trial_data
)

# Baseline conditioning (default) — T0 is efficiency, T1 and T2 are responsiveness
res_base <- randomRegress(
  model,
  Env  = "Treatment:Site:Variety",
  levs = c("T0", "T1", "T2"),
  type = "baseline"
)

# Sequential (Gram-Schmidt) — fully orthogonal components
res_seq <- randomRegress(
  model,
  Env  = "Treatment:Site:Variety",
  levs = c("T0", "T1", "T2"),
  type = "sequential"
)

# Partial — each treatment vs all others simultaneously
res_part <- randomRegress(
  model,
  Env  = "Treatment:Site:Variety",
  levs = c("T0", "T1", "T2"),
  type = "partial"
)

# Custom — T1 on T0 only; T2 on both T0 and T1
res_cust <- randomRegress(
  model,
  Env  = "Treatment:Site:Variety",
  levs = c("T0", "T1", "T2"),
  type = "custom",
  cond = list(T0 = NULL, T1 = "T0", T2 = c("T0", "T1"))
)

# Plot
plot_randomRegress(res_base, type = "quadrant")
plot_randomRegress(res_base, type = "gmat")
```

---

## Return value summary

`randomRegress()` returns a named list:

| Element | Type | Description |
|---------|------|-------------|
| `blups` | Data frame | Raw BLUPs + responsiveness BLUPs + HSD per variety × site |
| `beta` | Named list | Site-specific $\boldsymbol{\beta}_j$ matrices, one per conditioned treatment |
| `Gmat` | Matrix | Original G-matrix |
| `TGmat` | Matrix | Transformed G-matrix $\mathbf{T}\mathbf{G}\mathbf{T}^\top$ |
| `sigmat` | Matrix | Conditional genetic variances $\sigma^2_{j \mid A_j}$ per site |
| `tmat` | Matrix | Full transformation matrix $\mathbf{T}$ |
| `cond_list` | Named list | Resolved conditioning structure |
| `type` | Character | Conditioning scheme used |

---

## Summary

| Task | Argument |
|------|----------|
| Baseline conditioning | `type = "baseline"` (default) |
| Fully orthogonal (Gram-Schmidt) | `type = "sequential"` |
| Partial genetic variances | `type = "partial"` |
| Bespoke conditioning | `type = "custom", cond = list(...)` |
| Regression scatter plot | `plot_randomRegress(res, type = "regress")` |
| Efficiency–responsiveness plot | `plot_randomRegress(res, type = "quadrant")` |
| G-matrix heatmap | `plot_randomRegress(res, type = "gmat")` |
| Default archetype highlights | `highlight = "default"` |
| Named variety highlights | `highlight = c("Var01", "Var15")` |
| Suppress highlights | `highlight = NULL` |
| Restrict treatments shown | `treatments = "T1"` |
| Export plot data | `return_data = TRUE` |
| Absolute yield scale | `centre = TRUE` |

---

## References

Cullis, B.R., Smith, A.B. & Coombes, N.E. (2006). On the design of early
generation variety trials with correlated data. *Journal of Agricultural,
Biological, and Environmental Statistics*, **11**(4), 381–393.

Smith, A.B. & Cullis, B.R. (2018). Plant breeding selection tools built on
factor analytic mixed models for multi-environment trial data. *Euphytica*,
**214**, 143.

Mrode, R.A. (2014). *Linear Models for the Prediction of Animal Breeding
Values*, 3rd edn. CABI, Wallingford.
