---
title: "Multivariate Random Regression of Treatment BLUPs"
subtitle: "`randomRegress()` and `plot_randomRegress()`"
author: "biomAid package"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 9
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(biomAid)
library(ggplot2)
library(asreml)
library(ASExtras4)
```

---

## Overview

`randomRegress()` decomposes variety BLUPs from a multi-environment trial
into two biologically interpretable components:

- **Efficiency** — a variety's genetic value under the baseline (typically
  low-input) treatment.
- **Responsiveness (or Tolerance)** — a variety's capacity to genetically respond to the application of a treatment.

The decomposition is derived from the **G-matrix** (genetic
variance–covariance matrix across Treatment × Site combinations) using the
Gaussian conditional distribution. It generalises naturally from two
treatments to any number, with four built-in conditioning schemes.

`plot_randomRegress()` visualises the result as a regression scatter plot,
an efficiency–responsiveness quadrant plot, or a G-matrix heatmap.

---

## Mathematical Framework

### The G-matrix

Let $\hat{\mathbf{u}}_s = (\hat{u}_{1s}, \ldots, \hat{u}_{ks})^\top$ be the
vector of variety BLUPs for $k$ treatments at site $s$, stacked across all
sites. The genetic variance–covariance structure is described by the
**G-matrix**:

$$
\mathbf{G} = \operatorname{Var}\!\left(\hat{\mathbf{u}}\right),
$$

with $k \times k$ blocks $\mathbf{G}_{jl}$ giving the genetic covariance
between treatments $j$ and $l$ averaged across sites. In ASReml-R V4 this
is the genetic variance/covariance matrix for the Treatment-Site:Variety random term.

---

### Conditional decomposition

For treatment $j$ with conditioning set $A_j$ (a subset of the other
treatment labels), the **multivariate conditional normal** distribution gives
the regression of $\hat{u}_{js}$ on $\hat{\mathbf{u}}_{A_j,s}$:

$$
\hat{u}_{js} \mid \hat{\mathbf{u}}_{A_j,s}
  \sim \mathcal{N}\!\left(
    \boldsymbol{\beta}_j^\top \hat{\mathbf{u}}_{A_j,s},\;
    \sigma^2_{j \mid A_j}
  \right),
$$

where the **regression coefficients** and **conditional genetic variance** are

$$
\boldsymbol{\beta}_j = \mathbf{G}_{A_j A_j}^{-1}\, \mathbf{G}_{A_j j},
\qquad
\sigma^2_{j \mid A_j} = G_{jj} - \mathbf{G}_{j A_j}\,\boldsymbol{\beta}_j.
$$

The **responsiveness BLUP** for variety $v$ at site $s$ is the residual from
this regression:

$$
\tilde{u}_{js,v} = \hat{u}_{js,v} - \boldsymbol{\beta}_j^\top \hat{\mathbf{u}}_{A_j,s,v}.
$$

Responsiveness BLUPs are:

- **zero-mean** within each site (inherited from the BLUPs)
- **orthogonal** to the conditioning BLUPs in the G-matrix sense
- **genetically independent** of the efficiency signal, with variance
  $\sigma^2_{j \mid A_j}$

---

### The transformation matrix

Collecting all conditioning relationships across treatments and sites defines
a sparse lower-triangular transformation matrix $\mathbf{T}$ such that:

$$
\tilde{\mathbf{u}} = \mathbf{T}\hat{\mathbf{u}},
\qquad
\operatorname{Var}(\tilde{\mathbf{u}}) = \mathbf{T}\mathbf{G}\mathbf{T}^\top
  \equiv \mathbf{V}_T.
$$

The diagonal blocks of $\mathbf{V}_T$ are the conditional genetic variances
$\sigma^2_{j \mid A_j}$; off-diagonal blocks measure residual genetic
associations between components. For `type = "sequential"` the full
$\mathbf{V}_T$ is **diagonal** — a complete Gram-Schmidt
orthogonalisation of the BLUP space.

---

### Conditioning schemes

The four built-in schemes differ only in how $A_j$ is chosen:

| Scheme | $A_j$ for treatment $j$ | $\mathbf{V}_T$ structure | Best for |
|--------|------------------------|--------------------------|----------|
| `"baseline"` | $\{\text{levs}[1]\}$ only | Block-diagonal | Comparing all treatments to one baseline |
| `"sequential"` | $\{\text{levs}[1], \ldots, \text{levs}[j-1]\}$ | Diagonal ($LDL^\top$ Cholesky) | Fully orthogonal components |
| `"partial"` | All other treatments | Dense | Partial genetic variances |
| `"custom"` | User-specified | Depends | Bespoke experimental logic |

---

## Setup

### Simulated data and model

> **Requires:** ASReml-R V4 and ASExtras4. The vignette is pre-rendered so
> no licence is needed to read it; run the code chunks only if you have
> both packages installed.

We use `simTrialData()` to generate a balanced multi-environment trial with
20 varieties, 4 sites, and 3 nitrogen treatments (T0 = control, T1 = low N,
T2 = high N). A composite `TSite` factor (e.g. `"T0-Env1"`) is created
to serve as the Treatment × Site classification for the ASReml model.

```{r simulate}
#| cache: true
out <- simTrialData(
  nvar       = 20L,
  nsite      = 4L,
  treatments = c("T0", "T1", "T2"),
  nrep       = 3L,
  seed       = 42L,
  verbose    = FALSE
)
dat       <- out$data
levs      <- c("T0", "T1", "T2")
dat$TSite  <- interaction(dat$Treatment, dat$Site, sep = "-")
dat$Row    <- factor(dat$Row)
dat$Column <- factor(dat$Column)
```

We fit a **factor analytic model** with 2 factors for the
Treatment × Site genetic covariance structure. FA models are the standard
choice for multi-environment trials: they are parsimonious, converge well,
and capture the dominant axes of genotype × environment interaction.

```{r fit_model}
#| cache: true
#| results: hide
asreml.options(trace = FALSE)
model <- asreml(
  yield    ~ TSite,
  random   = ~ fa(TSite, 2):Variety + at(Site):Rep,
  residual = ~ dsum(~ ar1(Row):ar1(Column) | Site),
  data     = dat,
  maxit    = 30L
)
model <- update(model)
```

```{r model_summary}
vc <- summary(model)$varcomp
vc[grepl("TSite", rownames(vc)), ]
```

### Running `randomRegress()`

With the fitted model, `randomRegress()` extracts the G-matrix from the
FA variance parameters and computes BLUPs, regression coefficients, and
responsiveness components using baseline conditioning (T0 as efficiency,
T1 and T2 as responsiveness):

```{r run_rr}
#| cache: true
res <- randomRegress(
  model,
  Env  = "TSite:Variety",
  levs = levs,
  type = "baseline",
  sep  = "-"
)
```

The return value:

```{r show_structure}
str(res, max.level = 1)
```

---

## Regression plot — `type = "regress"`

The regression plot shows the raw treatment BLUPs for each conditioned
treatment (y-axis) against the conditioning treatment BLUPs (x-axis). Each
panel corresponds to one site × treatment pair. A **dotted random regression
line** through the origin carries the site-specific slope $\hat{\beta}$,
annotated in the top-left corner.

The slope $\hat{\beta}$ quantifies the average relationship between
treatments: $\hat{\beta} > 1$ means the conditioned treatment varies more
than the baseline; $\hat{\beta} < 1$ means it varies less.

```{r plot_regress}
#| fig-height: 7
p_reg <- plot_randomRegress(res, type = "regress")
print(p_reg)
```

ASReml BLUPs are already zero-mean within each site by construction, so
`centre = FALSE` (the default) is appropriate. Use `centre = TRUE` only
if BLUPs have been computed from adjusted means and retain a site-level
offset.

---

## Quadrant plot — `type = "quadrant"`

The quadrant plot separates the two components of genetic performance.
The **x-axis** carries the unconditional (efficiency) BLUPs; the **y-axis**
carries the responsiveness BLUPs $\tilde{u}_{js,v}$. Dotted zero lines on
both axes divide each panel into four quadrants:

| Quadrant | Efficiency | Responsiveness | Interpretation |
|----------|-----------|----------------|----------------|
| Top-right | High | High | Broadly adapted: performs well and responds strongly |
| Top-left | Low | High | Responsive but poor performance under baseline |
| Bottom-right | High | Low | Efficient but does not respond well to treatment |
| Bottom-left | Low | Low | Poor performance on both scales |

```{r plot_quadrant}
#| fig-height: 7
p_quad <- plot_randomRegress(res, type = "quadrant")
print(p_quad)
```

---

## Variety highlighting

By default, `highlight = "default"` automatically selects and annotates
**six varieties** that are informative archetypes of the efficiency–
responsiveness space. The algorithm operates in the centred quadrant space,
independently of the `centre` argument used for the plot axes.

### How the default algorithm works

1. Compute the mean efficiency (x) and responsiveness (y) for each variety
   **across all sites and treatment pairs**.
2. Separate varieties into the **top-right** quadrant ($x > 0$, $y > 0$)
   and the **bottom-left** quadrant ($x < 0$, $y < 0$).
3. Within each quadrant, retain varieties whose distance from the origin
   exceeds the 50th percentile (relaxes to 33rd then 0th if fewer than
   three candidates remain).
4. Pick three archetypes by polar angle $\theta = \arctan(|y|/|x|)$:
   - **Efficiency archetype** — smallest $\theta$ (close to the x-axis)
   - **Balanced archetype** — closest to $\theta = \pi/4$
   - **Responsiveness archetype** — largest $\theta$ (close to the y-axis)

The same six varieties are annotated consistently across all site and
treatment-pair panels.

```{r highlight_default}
#| fig-height: 7
# Orange = top-right archetypes, Blue = bottom-left archetypes
p_hl <- plot_randomRegress(res, type = "quadrant", highlight = "default")
print(p_hl)
```

### User-specified highlights

Pass a character vector of variety names to `highlight` to annotate specific
varieties of interest (shown in green):

```{r highlight_custom}
#| fig-height: 7
user_vars <- levels(factor(res$blups$Variety))[c(1L, 10L, 20L)]
p_custom  <- plot_randomRegress(res, type  = "quadrant",
                                highlight = user_vars)
print(p_custom)
```

### No highlighting

Set `highlight = NULL` to suppress annotation entirely and draw all points
in a single colour:

```{r highlight_none}
#| fig-height: 7
p_none <- plot_randomRegress(res, type = "quadrant", highlight = NULL)
print(p_none)
```

---

## Filtering treatments — `treatments` argument

For results with many conditioned treatments, `treatments` restricts which
pairs appear in the plot:

```{r treatments_filter}
#| fig-height: 4
p_t1 <- plot_randomRegress(res, type = "quadrant", treatments = "T1")
print(p_t1)
```

---

## G-matrix heatmap — `type = "gmat"`

The `"gmat"` plot displays `cov2cor(Gmat)` — the genetic correlations
between all Treatment × Site combinations — as a diverging heatmap. High
positive correlations (dark red) indicate treatments and/or sites that share
genetic rankings; negative correlations (dark blue) indicate rank reversal.

```{r plot_gmat}
#| fig-height: 6
#| fig-width: 8
p_gmat <- plot_randomRegress(res, type = "gmat")
print(p_gmat)
```

Correlations within the same treatment across sites (e.g. T0-Env1 with
T0-Env2) tend to be moderate to high. Correlations between treatments
(e.g. T0-Env1 with T2-Env3) reflect the degree of Treatment × Genotype ×
Environment interaction.

---

## Extracting plot data — `return_data = TRUE`

All three plot types support `return_data = TRUE`, which returns the tidy
data frame used to build the plot. This is useful for bespoke
visualisations or downstream analysis.

```{r return_data}
df_quad <- plot_randomRegress(res, type = "quadrant", return_data = TRUE)
head(df_quad)
```

The data frame can be passed to any `ggplot2` workflow:

```{r bespoke}
#| fig-height: 4
# Custom plot: quadrant for T1 only, custom colour by site
df_t1 <- plot_randomRegress(res, type = "quadrant",
                             treatments  = "T1",
                             return_data = TRUE)

ggplot(df_t1, aes(x = x, y = y, colour = Site)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_point(size = 1.8, alpha = 0.8) +
  facet_wrap(~ Site, nrow = 2L) +
  labs(x = "Efficiency (T0 BLUP)", y = "Responsiveness (T1 | T0)",
       title = "Custom site-coloured quadrant plot") +
  theme_bw() +
  theme(legend.position = "none")
```

---

## Return value summary

`randomRegress()` returns a named list:

| Element | Type | Description |
|---------|------|-------------|
| `blups` | Data frame | Raw BLUPs + responsiveness BLUPs + HSD per variety × site |
| `beta` | Named list | Site-specific $\boldsymbol{\beta}_j$ matrices, one per conditioned treatment |
| `Gmat` | Matrix | Original G-matrix |
| `TGmat` | Matrix | Transformed G-matrix $\mathbf{T}\mathbf{G}\mathbf{T}^\top$ |
| `sigmat` | Matrix | Conditional genetic variances $\sigma^2_{j \mid A_j}$ per site |
| `tmat` | Matrix | Full transformation matrix $\mathbf{T}$ |
| `cond_list` | Named list | Resolved conditioning structure |
| `type` | Character | Conditioning scheme used |

---

## Summary

| Task | Argument |
|------|----------|
| Baseline conditioning | `type = "baseline"` (default) |
| Fully orthogonal (Gram-Schmidt) | `type = "sequential"` |
| Partial genetic variances | `type = "partial"` |
| Bespoke conditioning | `type = "custom", cond = list(...)` |
| Regression scatter plot | `plot_randomRegress(res, type = "regress")` |
| Efficiency–responsiveness plot | `plot_randomRegress(res, type = "quadrant")` |
| G-matrix heatmap | `plot_randomRegress(res, type = "gmat")` |
| Default archetype highlights | `highlight = "default"` |
| Named variety highlights | `highlight = c("Var01", "Var15")` |
| Suppress highlights | `highlight = NULL` |
| Restrict treatments shown | `treatments = "T1"` |
| Export plot data | `return_data = TRUE` |
| Absolute yield scale | `centre = TRUE` |

---

## References

Lemerle, D., Smith, A., Verbeek, B., Koetz, E., Lockley, P. & Martin, P. (2006).
Incremental crop tolerance to weeds: A measure for selecting competitive ability
in Australian wheats. *Euphytica*, **149**, 85–95.
<https://doi.org/10.1007/s10681-005-9056-5>

McDonald G, Bovill W, Taylor J, Wheeler R. (2015) Responses to phosphorus among wheat genotypes.
*Crop and Pasture Science*, **66**(5), 430–444. (2015). <https://doi.org/10.1071/CP14191>

The WtmsDW Locus on Wheat Chromosome 2B Controls Major Natural Variation for
Floret Sterility Responses to Heat Stress at Booting Stage.
*Frontiers in Plant Science*, **12**, 635397. (2021).
<https://doi.org/10.3389/fpls.2021.635397>
